---
title: 'Lecture 4: Advanced Visualization & Tidy Data'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center")
library(tidyverse)
```

In [Lecture 2](lecture2.html), we saw that visualization using ggplot could provide important insights into large datasets.
Today, we will learn more about ggplot, reshape data for plotting, and build several more advanced visualizations.


## Visualizing the Diving Dataset

In this course, we will be using the package ggplot2 for all of our data visualization. The `gg` stands for "grammar of graphics", a framework for data visualization. This framework separates the process of visualization into different components: data, aesthetic mappings, and geometric objects. These components are then added together (or layered) to produce the final graph. We're going to illustrate these components using the [diving dataset](data/diving.csv) from Prof. Wyner's lecture.

```{r load diving, message=FALSE}
library(tidyverse)
diving <- read_csv("data/diving.csv")
```


### Components of a Plot

The first step in any data visualization is to tell R which tbl the data we want to plot lives. This is done using the `ggplot()` function. Notice that we are *assigning* the plot to a new variable, `diving_hist`. Later, we will *add* layers to the plot using the `+` operator.

```{r diving ggplot}
diving_hist <- ggplot(data = diving)
```

Aesthetics map the data to the properties of the plot. Examples include:

* `x`: the variable that will be on the x-axis
* `y`: the variable that will be on the y-axis
* `color`: the variable that categorizes data by color
* `shape`: the variable that categorizes data by shape

You can define the `aes` in the `ggplot` call, which will then be used for all later layers, or you can define the `aes` in the `geom` (see below), which will only apply to that `geom`. Geometric objects, or `geoms`, determine the type of plot that will be created. Examples include:

* `geom_point()`: creates a scatterplot
* `geom_histogram()`: creates a histogram
* `geom_line()`: creates a line
* `geom_boxplot()`: creates a boxplot


### Putting it All Together

Let's make a histogram of judges' scores.

```{r diving hist}
diving_hist <- 
  diving_hist + 
  geom_histogram(aes(x = JScore), binwidth = 0.25)
diving_hist
```

In the code above, we first overwrote `diving_hist` so that it is now a histogram of judges' scores where each bin has width 0.25. In the second line, we asked R to display this object. For ggplot code, place the `+` at the end of a line, just like the pipe `%>%`.

Notice that the label for the x-axis is JScore, which is the column name from the tbl. We can change the label by *adding* a layer.

```{r diving hist x label}
diving_hist <- 
  diving_hist + 
  labs(x = "Judge Score")
diving_hist
```


### Facets

What if we want to separate the judges' scores by country? We can use *facets*. Facets allow you to separate graphs by category. We do not need to redo our code for the histogram; we only need to add a *facet* layer to our graph `diving_hist`. The first argument of `facet_wrap()` is the column of our dataset that contains the category information. The second argument of `facet_wrap()` specifies the number of rows in which to display the graphs.

```{r facet_wrap}
diving_hist <- 
  diving_hist + 
  facet_wrap("JCountry", nrow = 3)
diving_hist
```

To get a sense of whether a particular country's judges are biased, it would be useful to add a reference line at the median score over all judges and countries to each facet. So, we must first calculate the overall median score, which we can do using the `summarize()` function. Then, we can display this median on our plots with `geom_vline()`, which adds a vertical line. We can also pass in a custom color argument to this line so that it will easily stand out from the rest of our values. This website provides a look at all the built-in color arguments in R: [R colors](https://sites.stat.columbia.edu/tzheng/files/Rcolor.pdf)

```{r facet_wrap median}
median_score <- diving %>% 
  reframe(med = median(JScore)) %>%
  pull(med)
diving_hist <- 
  diving_hist + 
  geom_vline(xintercept = median_score, color = "lightcoral")
diving_hist
```

Note that when calculating our `median_score`, the `reframe` function returns a tbl rather than a single value. To "pull" out the single median value from our tbl, we use the `pull` function.


### Boxplots and violin plots

Boxplots are also an important visualization tool. We now create boxplots of the judges' scores, separated by diving round. We can also color our boxplots according to the `Round` variable by including the `fill` argument within the `aes()` function. Note that if we included the `fill` argument outside of the `aes()` function, we would only be able to color all boxplots the same color (by putting `fill` inside of the aesthetic, we are able to "map" the variable `Round` to the fill color of the plot based on the values of `Round`).

```{r boxplot fill}
diving_box <- ggplot(data = diving) + 
  geom_boxplot(aes(x = Round, y = JScore, fill = Round)) +
  labs(title = "Diving scores by round", x = "", y = "Judge Score", fill = "Round")
diving_box
```

We can also remove the unnecessary x-axis ticks and labels as the legend on the right is sufficient. We do so using the `theme` layer:

```{r boxplot theme}
diving_box <- diving_box + 
  theme(axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank())
diving_box
```

We can also plot this data as violin plots, which take the same arguments as `geom_boxplot()` but use `geom_violin()` instead. We can also change the "look" of our plots by setting the theme. `theme_minimal()` is frequently used for a much cleaner look. 
```{r}
diving_violin <- ggplot(data = diving) + 
  geom_violin(aes(x = Round, y = JScore, fill = Round)) +
  labs(title = "Diving scores by round", x = "", y = "Judge Score", fill = "Round") +
  theme_minimal()

diving_violin
```

We can see that the boxplots provide more information about statistical features, such as the median and quartiles, but the violin plots provide a better sense of how the data is distributed. We can see the effect of smaller rounds having more binned scores, whereas they are spread much more evenly in the first round. 

### Barplots

We can also create barplots using `geom_bar`. By mapping `fill=JCountry` inside our `aes()`, we can map each bar's color to the country for that judge.

```{r barplot}
bar <- ggplot(data = diving) +
  geom_bar(aes(x = JCountry, fill = JCountry)) + 
  labs(x = "Judge Country") +
  theme_minimal() +
  theme(legend.position = "none") 
bar
```


### Scatterplots

Now let's turn back to scatterplots, which were introduced in [Lecture 2](lecture2.html). We plot judges' score versus rank of the diver. As we expect, divers with better (lower-numbered) ranks tend to receive higher scores.
```{r scatterplot}
scatter_raw = ggplot(data = diving) +
  geom_point(aes(x = Rank, y = JScore, color = Country)) +
  labs(x = "Diver Rank", y = "Judge Score") + 
  theme_minimal()

scatter_raw
```
We can see that this plot is a little messy and hard to interpret, which may commonly happen with scatterplots when multiple rows correspond to the same thing, such as a diver. We can improve on this visual by grouping per diver and using `reframe()` to calculate the mean of the judge scores, ranks, and difficulties for each diver. We also want to save the country for visualizations, so we use the `first()` command within `reframe()`. 

```{r group diving}
diving_grouped <- diving %>% 
  group_by(Diver) %>% 
  reframe(JScore_mean = mean(JScore), 
          Rank_mean = mean(Rank),
          Difficulty_mean = mean(Difficulty),
          Country = first(Country)) 

head(diving_grouped)  
```

This will allow us to plot a single point per diver, which will make the plot much easier to interpret.

```{r scatterplot grouped}

scatter = ggplot(data = diving_grouped) +
  geom_point(aes(x = Rank_mean, y = JScore_mean, color = Country), size = 2) +
  labs(x = "Diver Rank", y = "Judge Score") + 
  theme_minimal()

scatter
```
We can see here the much clearer negative relationship between judge score and diver rank.

### Lines

We can also add an `abline` to our scatterplot--that is, a line where we specify the y-intercept (`intercept`) and the slope (`slope`):

```{r scatter abline}
scatter <- scatter + 
  geom_abline(intercept = 8.5, slope = -0.1, color = "red")
scatter
```

*Note*: `geom_abline` is different from `geom_line`: `geom_line` "connects the dots" between your data and so doesn't have to be a straight line, whereas `geom_abline` draws a straight line with the specified slope and y-intercept. A common usage is to add a visual reference line to a scatterplot.

Other `geom` for lines are:

* `geom_vline`: to add a vertical line to a plot
* `geom_hline`: to add a horizontal line to a plot


### Stats

We can also specify a layer using `stat_`, which stands for statistical transformation. This is useful if we want to plot a summary statistic of our data, such as a mean or median. By using a `stat_` layer, we do not have to compute this summary statistic beforehand--`ggplot` will compute the summary statistic for us and then plot the result. 

For example, suppose we want to plot the means of each judge's score and provide error bars of one standard error on either side of the mean. We could use `summarize` and `group_by` to find the mean and standard errors for each judge, *or* we could just use a `stat_` layer!

The layer `stat_summary()` computes and then plots a user-specified summary statistic. We choose the option `mean_se` to calculate the means and standard errors of the scores of each judge.

As always, we set up the plot by calling `ggplot`, specifying `data = diving` and then providing the `aes`. In this case, we want the judge on the `x`-axis and their scores on the `y`-axis. We then add our `stat_summary` layer.

```{r stat_summary}
judges <- ggplot(data = diving, aes(x = Judge, y = JScore)) + 
  stat_summary(fun.data = mean_se) + 
  labs(y = "Judge Score")
judges
```

We can see that the judges' names are bunched together... we can make them much more readable by rotating the x-axis labels by 45 degrees using the `theme` layer:
```{r}
judges_rotate <- judges + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
judges_rotate
```

We can also completely flip the coordinates to make a horizontal plot!
```{r coord_flip}
judges <- judges + coord_flip()
judges
```

## Tidy Data for Visualization

Before we move to a few more advanced plots, it will be useful to reshape a dataset. Many plots become much easier once the data are in a tidy format:

1. Each **variable** must have its own **column**.
2. Each **observation** must have its own **row**.
3. Each **value** must have its own **cell**.

To make this concrete, let's go back to the dataset we created in the challenge question at the end of [Problem Set 3](ps3.html). If you haven't yet completed that, you can download the final [batting_2014_2015 dataset here](data/batting_2014_2015.RData) (then save it to your "data" folder). Printing out the tbl, we see that there is a separate row for each "player-season".

```{r load batting_2014_2015}
load("data/batting_2014_2015.RData")
batting_2014_2015
```

Let's say we want to study the relationship between a player's 2014 batting average and his 2015 batting average. In this case, each observation is a player, and that player has two associated variables: batting average in 2014 and batting average in 2015. We can do this with the `pivot_wider()` function. The `names_from` argument indicates the new column(s) in the table - in this case `yearID`. The `values_from` argument then indicates what values go in our new columns - in this case `BA`. We can use the `names_prefix` argument to rename the columns with `BA_` so that our column names do not start with numbers.

```{r pivot_wider batting_2014_2015}
batting_2014_2015 <- batting_2014_2015 %>%
  pivot_wider(
    names_from = yearID,
    values_from = BA,
    names_prefix = "BA_"
  )
head(batting_2014_2015)
```

It can also be helpful to work with the original, "long" format of the data when making visualizations. We can easily revert back to this form with `pivot_longer()`, which will take the compact version and unpack it back into one row per year. `cols` represents the column titles that are being converted into a new column, in this case `BA_2014` and `BA_2015`. The `names_to` argument indicates the name of the new column that will contain these values, and the `values_to` argument indicates the name of the new column that will contain the values themselves. We can then `mutate()` to remove the `BA_` prefix and see that our table looks like our original data.

```{r pivot_longer batting_2014_2015}
batting_2014_2015_long <- batting_2014_2015 %>%
  pivot_longer(
    cols = starts_with("BA_"),
    names_to = "yearID",
    values_to = "BA"
  ) %>% 
  mutate(yearID = sub("BA_", "", yearID))

head(batting_2014_2015_long)
```

We can use this `batting_2014_2015_long` tbl to make a bar plot with `geom_col()` which shows the batting averages of each player in 2014 and 2015.
```{r}
batting_2014_2015_slice <- batting_2014_2015_long %>%
  slice_head(n = 20) # take the first 20 players alphabetically 

ggplot(batting_2014_2015_slice, aes(x = playerID, y = BA, fill = yearID)) +
  geom_col(position = "dodge", color = "black", alpha = 0.8) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Heat Maps and Plot Annotations

When both axes are quantitative, it can be useful to divide the plane into rectangular bins and summarize a third variable within each bin. The function `stat_summary_2d()` does that automatically.

For instance, we can look at how batting average changed from 2014 to 2015 by first creating a column for the year-to-year change.

```{r batting change}
batting_2014_2015 <- batting_2014_2015 %>%
  mutate(BA_change = BA_2015 - BA_2014)
```

Then we can compute the mean change within rectangular bins of 2014 and 2015 batting averages. The function `scale_fill_distiller()` lets us control the fill colors and legend title, and `annotate()` lets us draw reference shapes on top of the plot.

```{r stat_summary_2d batting}
ggplot(batting_2014_2015, aes(x = BA_2014, y = BA_2015, z = BA_change)) +
  stat_summary_2d() +
  scale_fill_distiller("Mean change", palette = "RdBu") +
  annotate("rect", xmin = 0.25, xmax = 0.30, ymin = 0.25, ymax = 0.30,
           alpha = 0, color = "black") +
  labs(x = "2014 Batting Average", y = "2015 Batting Average")
```

If your data are already summarized into a rectangular grid, `geom_tile()` is a direct way to build a heat map.

```{r geom_tile}
diving_tile <- diving %>%
  group_by(Round, Country) %>%
  reframe(mean_score = mean(JScore))

ggplot(diving_tile, aes(x = Round, y = Country, fill = mean_score)) +
  geom_tile() +
  labs(x = "Round", y = "Country", fill = "Mean Score")
```

## Scales and 2D Histograms

We already used `scale_fill_distiller()` to control the fill colors in a heat map. We can also use scales to control point colors. We return to the scatter plot of the judges' scores vs rank of the divers. This time, we want to color the points by the difficulty of the dive.

We use the layer `scale_color_distiller()`. The second word, `color`, is the `aes` we want to change. We can replace it with `x`, `y` or `fill`, depending on the `aes` we want to change. 

The third word is `distiller`, which we use because our `color` variable, `Difficulty`, is continuous. If it were discrete, we would write `brewer` instead.

```{r scales scatter}
scatter <- ggplot(data = diving_grouped) +
  geom_point(aes(x = Rank_mean, y = JScore_mean, color = Difficulty_mean)) +
  scale_color_distiller(palette = "OrRd", direction = 1) 
  
scatter
```

Interestingly, it seems that some of the highest ranked divers perform most of the less difficult dives, but perform these easy dives very well. 

To further investigate, we plot a 2D histogram of Rank vs Difficulty.

```{r scales hist}
hist <- ggplot(data = diving)  + 
  geom_bin2d(aes(x = Rank, y = Difficulty), bins = 10) + 
  scale_fill_distiller(palette = "Spectral")
hist
```

Note that we use `fill` instead of `color` in `scale_fill_distiller()` to control the fill of the histogram bins.

From the 2D histogram, we can see that the higher ranked divers attempt both more difficult and less difficult dives, unlike the lower ranked divers who only attempt more difficult dives. 
