---
title: 'Lecture 6: Linear Regression'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center")
library(tidyverse)
```

In [Lecture 5](lecture5.html), we predicted an MLB player's 2015 batting average using their 2014 batting average. Binning helped us see the relationship, but it required us to choose arbitrary bin widths. In this lecture, we will use *correlation* to measure the strength of the linear relationship and *linear regression* to fit a smooth prediction line without choosing bins.

We again load the data (if you don't have it, you can download it [here](data/batting_2014_2015.RData)) and use the `pivot_wider()` function to separate the 2014 and 2015 batting averages into separate columns. 

```{r load batting_2014_2015, message = FALSE, cache = FALSE}
load("data/batting_2014_2015.RData")
batting_2014_2015 <- batting_2014_2015 %>%
  pivot_wider(
    names_from = yearID,
    values_from = BA,
    names_prefix = "BA_"
  )
head(batting_2014_2015)
```

Again, we plot the 2014 batting averages against the 2015 batting averages:

```{r plot batting_2014_2015}
g <- ggplot(data = batting_2014_2015) +
  geom_point(aes(x = BA_2014, y = BA_2015)) + 
  labs(x = "2014 Batting Average", y = "2015 Batting Average")
g
```


## Correlation

We can see a positive trend between the 2014 and 2015 batting averages. How can we quantify the relationship between the 2014 and 2015 data? 

The **correlation coefficient**, *r*, is one way to summarize the dependence between two seasons with one number. *r* is a standardized measure of the *linear* dependence between two variables (usually called $x$ and $y$) and can take values between -1 and +1.  

If $r = 1$, then the points all lie on a line with positive slope. If $r = -1$, the points all lie on a line with negative slope. 

Loosely speaking, we can think of correlation as a measure of how well a line fits our data. [Let's get some practice!](https://guessthecorrelation.com/)

We can calculate the correlation between two variables in `R` using the function `cor()`. Since `cor()` expects vectors rather than a tibble, we cannot pipe our data into it directly. Instead, we need to specifically point to each variable within the dataset, using the dollar sign symbol we've seen a couple times before: `our_dataset$our_variable`. This extracts a single column as a vector. By contrast, `our_dataset %>% select(our_variable)` returns a one-column tibble, which is useful in other contexts but is not what we want here.

The correlation between the 2014 and 2015 batting averages is:

```{r cor}
cor(batting_2014_2015$BA_2014, batting_2014_2015$BA_2015)
```

We have a moderate amount of correlation between 2014 and 2015 batting averages, but there is still a lot of noise. 

It is important to remember that correlation is only a measure of *linear* dependence between two variables. The below example shows data where $y = x^2$ exactly. 

```{r quadratic plot, echo = FALSE}
set.seed(1234)
x <- runif(50, -1, 1)
y <- x^2
dat <- tibble(x = x, y = y)
r_quadratic <- cor(x, y)
eg <- ggplot(data = dat, aes(x, y)) +
  geom_point() +
  labs(title = expression(paste("y = ", x^2)))
eg
```

Even though $y$ is determined exactly by $x$, $r = `r round(r_quadratic, 2)`$, indicating a weak *linear* relationship. A straight line fit to these data will therefore be nearly flat.

```{r quadratic plot line, echo = FALSE}
quadratic_line <- lm(y ~ x, data = dat)
eg <- eg +
  geom_abline(
    intercept = coef(quadratic_line)[1],
    slope = coef(quadratic_line)[2],
    color = "red"
  )
eg
```

```{r slope intercept, echo=FALSE}
best_fit <- lm(BA_2015 ~ BA_2014, data = batting_2014_2015)
a_best <- best_fit$coefficients[1]
b_best <- best_fit$coefficients[2]
x_example <- 0.31
y_example <- predict(best_fit, newdata = tibble(BA_2014 = x_example))[[1]]
```

Now let's go back to the 2014 and 2015 batting data. To visualize their linear relationship, we draw the *line of best fit*. In this example, the line has a $y$-intercept of `r round(a_best, 3)` and a slope of `r round(b_best, 3)`. We will see how to estimate these values below.

```{r slope intercept plot}
g <- g + geom_abline(intercept = a_best, slope = b_best, color = "red") + 
  labs(title = paste0("y = ", round(a_best, 3), " + ", round(b_best, 3), "x"))
g
```

**Question:** Which feature of the best fit line, $a$ or $b$, quantifies the linear dependence between 2014 and 2015 batting average?

How does correlation relate to our line of best fit? This brings us to the *regression method*.


## The Regression Method

If we standardize both variables, $x$ and $y$, the correlation is the *slope* of the line of best fit:

$$\frac{y - \bar{y}}{sd(y)} = r \times \frac{x - \bar{x}}{sd(x)}$$

We can unpack this equation to get the formula for our unstandardized line of best fit:

$$y = a + bx$$
where $a = \bar{y} - b\bar{x}$ and $b = r \times sd(y)/sd(x)$.

Now that we have our regression line, we can predict $y$ when we know $x$. For example, if a player's 2014 batting average was `r x_example`, their predicted 2015 batting average is:
$$\widehat{y} = `r round(a_best, 3)` + `r round(b_best, 3)` \times `r x_example` = `r round(y_example, 3)`$$
We use $\widehat{y}$ for a predicted value of $y$. The regression line estimates the average value of $Y$ for players with a given value of $x$.

```{r plot predicted y, echo = FALSE, warning = FALSE}
g <- ggplot(data = batting_2014_2015) + 
  geom_point(aes(x = BA_2014, y = BA_2015)) + 
  labs(x = "2014 Batting Average", y = "2015 Batting Average") + 
  geom_abline(intercept = a_best, slope = b_best, color = "red") +
  labs(title = paste0("y = ", round(a_best, 3), " + ", round(b_best, 3), "x")) +
  annotate(
    "segment",
    x = 0.19, xend = x_example,
    y = y_example, yend = y_example,
    linetype = "dashed", color = "dodgerblue"
  ) +
  annotate(
    "segment",
    x = x_example, xend = x_example,
    y = 0.20, yend = y_example,
    linetype = "dashed", color = "dodgerblue"
  ) +
  annotate("point", x = x_example, y = y_example, color = "dodgerblue", size = 2) +
  coord_cartesian(xlim = c(0.19, 0.355), ylim = c(0.20, 0.35), expand = FALSE)
g
```

**Question:** What is the interpretation of the slope coefficient, $b$?


## Fitting Linear Models

To find the regression coefficients, $a$ and $b$, we use the function `lm()`. The usage of `lm()` is as follows:

```{r lm fn}
model <- lm(BA_2015 ~ BA_2014, data = batting_2014_2015)
```

The first argument in the `lm()` call is called a `formula`. It takes the form `y ~ x`, where `y` is our response and `x` is our predictor. In the `lm()` call above, `BA_2015` is our response and `BA_2014` is our predictor.

The second argument in `lm()` is where we specify our data: `batting_2014_2015`. `R` then looks for the columns `BA_2014` and `BA_2015` in the given dataset to calculate the regression coefficients.

Our new object, `model`, contains a lot of information about our regression line. At the moment, we just want the coefficients. We can access the coefficients as follows:
```{r get coefficients}
model$coefficients
```

We could plot the fitted line by passing these coefficients to `geom_abline()`. A more general approach is to use `predict()` to add fitted values to our dataset, sort the rows by the predictor, and plot those values with `geom_line()`:
```{r plot predictions, warning = FALSE }
batting_2014_2015 <- batting_2014_2015 %>%
  mutate(pred = predict(model))

ggplot(data = arrange(batting_2014_2015, BA_2014)) +
  geom_point(aes(x = BA_2014, y = BA_2015)) + 
  geom_line(aes(x = BA_2014, y = pred), color = "lightcoral", linewidth = 1) +
  labs(x = "2014 Batting Average", y = "2015 Batting Average") +
  labs(title = "Batting Average with Regression Line")
```

## Measuring Fit and Calculating Residuals

For a linear regression with one predictor, $R^2$ is the square of the correlation, $r^2$. It measures the proportion of variation in the response explained by the model. We can retrieve it from the model summary:

```{r rsquare}
summary(model)$r.squared
```

The *residual* for a row is its observed outcome minus its predicted outcome. Because we already have the predictions in `pred`, we can calculate residuals directly with `mutate()`:

```{r add predictions residuals}
batting_2014_2015 <- batting_2014_2015 %>% 
  mutate(
    pred = predict(model),
    resid = BA_2015 - pred
  )
batting_2014_2015
```

We can display each residual as the vertical distance between an observed value and the regression line:

```{r plot residuals}
g <- ggplot(batting_2014_2015) +
  geom_segment(aes(x = BA_2014, xend = BA_2014, y = BA_2015, yend = pred), color = "dodgerblue") + 
  geom_point(aes(x = BA_2014, y = BA_2015)) + 
  geom_abline(intercept = a_best, slope = b_best, color = "red") +
  labs(x = "2014 Batting Average", y = "2015 Batting Average")
g
```


## Predicting New Observations

So far, we have looked at fitted values for the *training* data: the data used to fit our linear model. Suppose we have more players' batting averages from 2014 but do not have their batting averages from 2015. We can use `predict()` to obtain predictions for these new players.

Below, we enter the players' 2014 batting averages in a tibble called `new_data`. The first argument of `predict()` is the fitted model. The `newdata` argument supplies the rows for which we want predictions.

```{r new_data predict}
new_data <- tibble(BA_2014 = c(0.241, 0.31, 0.265))
new_pred <- predict(model, newdata = new_data)
new_pred
```

We can add those predictions to `new_data` with `mutate()`:

```{r new_data add predictions}
new_data <- new_data %>%
  mutate(pred = predict(model, newdata = new_data))
new_data
```

We add these predictions to our plot:

```{r plot new predictions}
ggplot(arrange(batting_2014_2015, BA_2014)) +
  geom_point(aes(x = BA_2014, y = BA_2015)) +
  geom_line(aes(x = BA_2014, y = pred), color = "lightcoral", linewidth = 1) +
  geom_point(data = new_data, mapping = aes(x = BA_2014, y = pred), color = "dodgerblue") +
  labs(x = "2014 Batting Average", y = "2015 Batting Average")
```
