---
title: 'Lecture 7: Transformations and Multiple Regression'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.align = "center")
library(tidyverse)
options(pillar.sigfig = 4, width = 100)
set.seed(2026)
```

Linear regression with one untransformed predictor fits a straight line. Transforming a predictor changes the fitted shape, while adding predictors produces a multiple regression. All of these models use `lm()`.

## Statcast Fly Ball Data {#statcastflyballdata}

The files [`statcast_models_train.csv`](data/statcast_models_train.csv) and [`statcast_models_test.csv`](data/statcast_models_test.csv) contain a subset of fly balls from the 2019 MLB regular season with complete Statcast measurements. The subset includes exit velocities from 66 to 111 miles per hour and launch angles from 20 to 60 degrees. These limits remove a small number of extreme measurements that would stretch the plot axes. The training data cover March through July, and the testing data cover August and September.

Each row is one fly ball. The models use three variables:

- `exit_velocity`: speed off the bat, in miles per hour
- `launch_angle`: vertical angle off the bat, in degrees
- `hit_distance`: projected distance traveled, in feet

All six candidate models are fit with the training data. The testing data remain unopened until the candidates are fixed, then testing RMSE measures how well the fitted relationships carry forward to later observations.

```{r load Statcast training data, message=FALSE}
statcast_train <- read_csv(
  "data/statcast_models_train.csv",
  show_col_types = FALSE
)
```

The plots use a random sample so that the point cloud remains visible. The model fits and RMSE calculations use every training fly ball.

```{r sample Statcast data for plots}
statcast_plot <- statcast_train %>%
  slice_sample(n = 10000)
```

```{r plot exit velocity and distance}
ggplot(statcast_plot, aes(x = exit_velocity, y = hit_distance)) +
  geom_point(alpha = 0.15) +
  labs(x = "Exit Velocity (mph)", y = "Hit Distance (feet)")
```

Fly balls hit harder generally travel farther, but a straight line is only one possible shape.

## Exit Velocity Transformations {#exitvelocitytransformations}

The linear model assumes that one additional mile per hour produces the same fitted difference at 70 mph as at 100 mph. The logarithmic and quadratic models allow that difference to change.

```{r fit exit velocity candidate models}
lin_velo_model <- lm(
  hit_distance ~ exit_velocity,
  data = statcast_train
)

log_velo_model <- lm(
  hit_distance ~ log(exit_velocity),
  data = statcast_train
)

quad_velo_model <- lm(
  hit_distance ~ exit_velocity + I(exit_velocity^2),
  data = statcast_train
)
```

The logarithm compresses larger exit velocities more than smaller ones, so its fitted curve flattens as exit velocity increases. The quadratic model allows the slope to increase or decrease.

The linear, logarithmic, and quadratic forms are `a + b * x`, `a + b * log(x)`, and `a + b1 * x + b2 * x^2`. Inside a model formula, `I()` tells R to perform ordinary arithmetic.

## Comparing the Exit Velocity Curves {#comparingexitvelocitycurves}

The same evenly spaced prediction grid is used for every model, so differences among the curves reflect the fitted shapes.

```{r create exit velocity prediction grid}
speed_grid <- tibble(
  exit_velocity = seq(
    min(statcast_train$exit_velocity),
    max(statcast_train$exit_velocity),
    by = 0.5
  )
) %>%
  mutate(
    lin_velo = predict(lin_velo_model, newdata = .),
    log_velo = predict(log_velo_model, newdata = .),
    quad_velo = predict(quad_velo_model, newdata = .)
  ) %>%
  pivot_longer(
    cols = c(lin_velo, log_velo, quad_velo),
    names_to = "model",
    values_to = "fitted_distance"
  ) %>%
  mutate(
    model = factor(
      model,
      levels = c("lin_velo", "log_velo", "quad_velo"),
      labels = c("Lin. velo.", "Log velo.", "Quad. velo.")
    )
  )
```

`pivot_longer()` places the three sets of predictions in one column and records their source in `model`. Mapping `model` to color draws the three curves together and creates the legend automatically.

```{r plot exit velocity candidate curves}
ggplot() +
  geom_point(
    data = statcast_plot,
    mapping = aes(x = exit_velocity, y = hit_distance),
    color = "gray40",
    alpha = 0.1,
    inherit.aes = FALSE
  ) +
  geom_line(
    data = speed_grid,
    mapping = aes(
      x = exit_velocity,
      y = fitted_distance,
      color = model
    ),
    linewidth = 1
  ) +
  labs(
    x = "Exit Velocity (mph)",
    y = "Hit Distance (feet)",
    color = NULL
  )
```

The linear curve has a constant slope, the logarithmic curve flattens, and the quadratic curve becomes steeper.

RMSE measures the typical prediction error in feet. Lower values are better. Each model's fitted values are stored in a column, then reshaped so that `reframe()` can apply the same RMSE calculation to every model.

```{r add exit velocity training predictions}
statcast_train <- statcast_train %>%
  mutate(
    lin_velo = predict(lin_velo_model),
    log_velo = predict(log_velo_model),
    quad_velo = predict(quad_velo_model)
  )
```

```{r exit velocity training RMSE}
rmse <- function(predicted, actual) {
  sqrt(mean((actual - predicted)^2))
}

speed_training_scores <- statcast_train %>%
  pivot_longer(
    cols = c(lin_velo, log_velo, quad_velo),
    names_to = "model",
    values_to = "prediction"
  ) %>%
  reframe(
    rmse = rmse(prediction, hit_distance),
    .by = model
  )

speed_training_scores
```

Within `reframe()`, `.by = model` calculates one RMSE for each model.

`lm()` chooses coefficients to reduce error on the training data. The quadratic model can reproduce the linear model by setting the coefficient on the squared term to zero, so its training RMSE cannot be larger. Testing data determine whether that added flexibility improves predictions on later observations.

## Adding Launch Angle

The residual from `lin_velo_model` is observed hit distance minus fitted hit distance. Positive residuals mark fly balls that traveled farther than fitted; negative residuals mark fly balls that traveled less far. Plotting residuals against launch angle shows whether angle explains errors left by the model using only exit velocity.

```{r sample exit velocity residuals}
statcast_residual_plot <- statcast_train %>%
  mutate(residual_distance = hit_distance - lin_velo) %>%
  slice_sample(n = 10000)
```

```{r plot exit velocity residuals against launch angle}
ggplot(
  statcast_residual_plot,
  aes(x = launch_angle, y = residual_distance)
) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_point(alpha = 0.15) +
  geom_smooth(
    method = "lm",
    formula = y ~ x + I(x^2),
    se = FALSE,
    color = "lightcoral",
    linewidth = 1
  ) +
  labs(
    x = "Launch Angle (degrees)",
    y = "Residual Hit Distance (feet)"
  )
```

The residuals rise and then fall. The model using only exit velocity underpredicts many fly balls near the middle of the angle range and overpredicts many fly balls with high launch angles.

Representing this rise and fall requires `launch_angle` together with `I(launch_angle^2)`. The squared term follows directly from the residual pattern.

The same quadratic launch angle specification is paired with each exit velocity shape. Holding the launch angle terms fixed isolates the consequence of the exit velocity transformation.

```{r fit multiple regression models}
lin_velo_quad_angle_model <- lm(
  hit_distance ~ exit_velocity + launch_angle + I(launch_angle^2),
  data = statcast_train
)

log_velo_quad_angle_model <- lm(
  hit_distance ~ log(exit_velocity) + launch_angle + I(launch_angle^2),
  data = statcast_train
)

quad_velo_quad_angle_model <- lm(
  hit_distance ~ exit_velocity + I(exit_velocity^2) +
    launch_angle + I(launch_angle^2),
  data = statcast_train
)
```

These are multiple regressions because each model uses exit velocity and launch angle. The exit velocity terms compare fly balls with the same launch angle. The two launch angle terms allow fitted distance to rise and then fall as launch angle increases.

```{r add multiple regression training predictions}
statcast_train <- statcast_train %>%
  mutate(
    lin_velo_quad_angle = predict(
      lin_velo_quad_angle_model
    ),
    log_velo_quad_angle = predict(
      log_velo_quad_angle_model
    ),
    quad_velo_quad_angle = predict(
      quad_velo_quad_angle_model
    )
  )
```

Each multiple regression contains its counterpart using only exit velocity when both launch angle coefficients are zero. Adding launch angle therefore cannot increase training RMSE. Testing RMSE determines whether the improvement carries forward.

## Selecting a Model with Testing Data

The six candidates are fixed: three use exit velocity alone, and three add quadratic launch angle. They are applied to the later months without refitting, so every testing prediction comes from a relationship learned through July.

```{r load Statcast testing data, message=FALSE}
statcast_test <- read_csv(
  "data/statcast_models_test.csv",
  show_col_types = FALSE
) %>%
  mutate(
    lin_velo = predict(lin_velo_model, newdata = .),
    log_velo = predict(log_velo_model, newdata = .),
    quad_velo = predict(quad_velo_model, newdata = .),
    lin_velo_quad_angle = predict(
      lin_velo_quad_angle_model,
      newdata = .
    ),
    log_velo_quad_angle = predict(
      log_velo_quad_angle_model,
      newdata = .
    ),
    quad_velo_quad_angle = predict(
      quad_velo_quad_angle_model,
      newdata = .
    )
  )
```

Stacking the training and testing predictions allows the same RMSE calculation for both samples. After the predictions are reshaped, `.by = c(sample, model)` produces one score for each combination of sample and model.

```{r compare model RMSE}
model_comparison <- bind_rows(
  Training = statcast_train,
  Testing = statcast_test,
  .id = "sample"
) %>%
  pivot_longer(
    cols = c(
      lin_velo, log_velo, quad_velo,
      lin_velo_quad_angle, log_velo_quad_angle,
      quad_velo_quad_angle
    ),
    names_to = "model",
    values_to = "prediction"
  ) %>%
  reframe(
    rmse = rmse(prediction, hit_distance),
    .by = c(sample, model)
  ) %>%
  mutate(
    sample = factor(sample, levels = c("Training", "Testing")),
    model = factor(
      model,
      levels = c(
        "lin_velo",
        "log_velo",
        "quad_velo",
        "lin_velo_quad_angle",
        "log_velo_quad_angle",
        "quad_velo_quad_angle"
      ),
      labels = c(
        "Lin. velo.",
        "Log velo.",
        "Quad. velo.",
        "Lin. velo. + quad. angle",
        "Log velo. + quad. angle",
        "Quad. velo. + quad. angle"
      )
    )
  )

model_comparison

ggplot(
  model_comparison,
  aes(x = model, y = rmse, fill = sample)
) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = scales::number(rmse, accuracy = 0.01)),
    position = position_dodge(width = 0.9),
    vjust = -0.4,
    size = 3
  ) +
  scale_x_discrete(
    labels = scales::label_wrap(18),
    guide = guide_axis(n.dodge = 2)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = NULL, y = "RMSE (feet)", fill = NULL)
```

Training RMSE measures fit to the observations used to estimate a model. Testing RMSE measures errors on later observations and provides the basis for selection.

Adding the curved launch angle relationship reduces testing RMSE from more than 30 feet to less than 15.3 feet. Launch angle explains substantial, repeatable information left by the models using only exit velocity.

Once launch angle enters the model, the exit velocity shape matters little. The quadratic velocity model has the smallest testing RMSE, but it beats the linear velocity model by only about 0.015 feet. The large improvement comes from adding launch angle, not from choosing a more complicated exit velocity shape.
