---
title: 'Lecture 6: Linear Regression'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.align = "center")
library(tidyverse)
```

In [Lecture 5](lecture5.html), we used bins to summarize how players' batting averages changed from 2014 to 2015. The bin averages described the pattern in the data. Linear regression gives us a single smooth line for that pattern, without requiring us to choose bin widths.

Load the data (you can download it [here](data/batting_2014_2015.RData)) and reshape it so that each player has separate columns for 2014 and 2015.

```{r load batting_2014_2015, message=FALSE}
load("data/batting_2014_2015.RData")

batting_2014_2015 <- batting_2014_2015 %>%
  pivot_wider(
    names_from = yearID,
    values_from = BA,
    names_prefix = "BA_"
  )

head(batting_2014_2015)
```

```{r plot batting_2014_2015}
ggplot(batting_2014_2015, aes(x = BA_2014, y = BA_2015)) +
  geom_point() +
  labs(x = "2014 Batting Average", y = "2015 Batting Average")
```

## Correlation

The **correlation coefficient**, *r*, summarizes the direction and strength of a linear relationship. It ranges from -1 to 1.

- A correlation near 1 describes a strong positive linear relationship.
- A correlation near -1 describes a strong negative linear relationship.
- A correlation near 0 describes a weak linear relationship.

We calculate a correlation with `cor()`. The function expects two vectors, so we use `$` to extract each column.

```{r correlation}
cor(batting_2014_2015$BA_2014, batting_2014_2015$BA_2015)
```

There is a positive relationship between the two seasons, but there is also considerable variation among players with similar 2014 batting averages.

Correlation measures only *linear* dependence. In the data below, `y` is determined exactly by `x`, but the relationship is curved and the correlation is near zero.

```{r quadratic plot, echo=FALSE}
set.seed(1234)
x <- runif(50, -1, 1)
quadratic_data <- tibble(x = x, y = x^2)

ggplot(quadratic_data, aes(x, y)) +
  geom_point() +
  labs(
    title = paste("Correlation =", round(cor(quadratic_data$x, quadratic_data$y), 2)),
    subtitle = "A strong relationship does not have to be linear"
  )
```

## Fitting a Linear Model

Use `lm()` to fit the line relating 2014 batting average to 2015 batting average.

```{r fit model}
batting_model <- lm(BA_2015 ~ BA_2014, data = batting_2014_2015)
batting_model$coefficients
```

The formula has the form `response ~ predictor`. Here, `BA_2015` is the response on the vertical axis and `BA_2014` is the predictor on the horizontal axis.

A fitted line has an intercept and a slope:

$$\widehat{y} = a + bx$$

The intercept, *a*, is the fitted value when the predictor is zero. That is far outside the batting averages in these data, so it is not substantively useful here.

The slope, *b*, is the fitted change in 2015 batting average associated with a one-unit increase in 2014 batting average. Since a one-unit change is enormous for batting average, it is clearer to interpret a 0.010 increase.

```{r slope interpretation}
coef(batting_model)["BA_2014"] * 0.010
```

**Question:** According to the model, how much higher is the fitted 2015 batting average when the 2014 batting average is 10 points higher?

## Fitted Values

`predict()` returns the fitted value for every row used to estimate the model. Add those values to the data and draw the line.

```{r add fitted values}
batting_2014_2015 <- batting_2014_2015 %>%
  mutate(fitted = predict(batting_model))

ggplot(arrange(batting_2014_2015, BA_2014)) +
  geom_point(aes(x = BA_2014, y = BA_2015)) +
  geom_line(
    aes(x = BA_2014, y = fitted),
    color = "lightcoral",
    linewidth = 1
  ) +
  labs(
    x = "2014 Batting Average",
    y = "2015 Batting Average",
    title = "Batting Average with Fitted Regression Line"
  )
```

These are fitted values for players whose 2015 batting averages are already in the data. They summarize the average relationship between the two seasons; they are not out-of-sample forecasts.

For example, the fitted value for a 2014 batting average of 0.310 is:

```{r fitted example}
predict(batting_model, newdata = tibble(BA_2014 = 0.310))
```

## Residuals and Model Fit

A **residual** is the observed value minus the fitted value. A positive residual means the observed 2015 batting average is above the line; a negative residual means it is below the line.

```{r add residuals}
batting_2014_2015 <- batting_2014_2015 %>%
  mutate(residual = BA_2015 - fitted)

head(batting_2014_2015)
```

```{r plot residuals}
ggplot(batting_2014_2015, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, color = "lightcoral", linewidth = 1) +
  geom_point() +
  labs(x = "Fitted 2015 Batting Average", y = "Residual")
```

A useful residual plot has points scattered around zero without a clear pattern. A curve or funnel shape would tell us that a straight line or constant spread does not describe the data well.

Two numbers summarize different aspects of model fit:

- $R^2$ is the proportion of variation in the response explained by the fitted line.
- RMSE is the typical size of a residual, in the same units as the response.

```{r fit statistics}
summary(batting_model)$r.squared

batting_2014_2015 %>%
  summarize(RMSE = sqrt(mean(residual^2)))
```

For a linear regression with one predictor, $R^2$ equals the squared correlation. A larger $R^2$ means the line explains more of the variation. A smaller RMSE means the fitted values are closer to the observed values.

## Predicting New Rows

Supply a new data frame to the `newdata` argument when the response is not already observed. Its predictor column must have the same name used in the model.

```{r predict new rows}
new_players <- tibble(BA_2014 = c(0.241, 0.310, 0.265)) %>%
  mutate(predicted_BA_2015 = predict(batting_model, newdata = .))

new_players
```

These predictions still describe the fitted relationship learned from the 2014--2015 data. Lecture 7 will separate the data used to fit a model from the data used to evaluate its predictions.

## How Correlation Determines the Line

If both variables are standardized, the correlation is the slope of the fitted line:

$$
\frac{y - \bar{y}}{sd(y)} = r \times \frac{x - \bar{x}}{sd(x)}
$$

Rearranging gives the unstandardized slope and intercept:

$$
b = r \times \frac{sd(y)}{sd(x)}, \qquad a = \bar{y} - b\bar{x}
$$

This explains why the sign of the slope always matches the sign of the correlation. In practice, `lm()` calculates the coefficients for us.
