---
title: 'Lecture 7: More About Regression'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center")
library(tidyverse)
options(pillar.sigfig = 4, width = 100)
```

In [Lecture 5](lecture5.html) and [Lecture 6](lecture6.html), we predicted a baseball player's 2015 batting average using their 2014 batting average. We found that some models fit the observed data well but might not predict new observations well. This is called *overfitting*: the model has learned random variation in the training data as though it were a repeatable pattern. We will now use NFL field-goal data to learn how a testing set can reveal overfitting.


## Training and Testing Paradigm

Suppose we have fit multiple models to a given dataset. How should we choose which model is the best?  

One strategy is to see how well each model predicts the data used to fit it. However, a flexible model can "over-learn" patterns in this *training set* that will not repeat in new data.

A common alternative is to split the original dataset into two parts: a *training set* and a *testing set*. We fit our models using only the training set, then evaluate their predictions on the testing set. Both sets are samples; the testing set acts as a stand-in for future, out-of-sample observations.


## Field Goal Success in the NFL

The file [nfl_fg_train.csv](data/nfl_fg_train.csv) contains data on NFL field goals attempted between 2005 and 2015. We will train several models of field-goal success with these data, then evaluate their predictions using the held-out attempts in [nfl_fg_test.csv](data/nfl_fg_test.csv). Download both files and move them to your `data` folder.

```{r load fg_train, message = FALSE}
fg_train <- read_csv("data/nfl_fg_train.csv")
```

The simplest forecast of field-goal success is the overall success rate. This forecast does not differentiate between kickers or account for distance or game context. The code below adds this forecast to the tibble in a column called `phat_all`.

```{r phat_all}
fg_train <- 
  fg_train %>%
  mutate(phat_all = mean(Success))
```

Kickers make just over 83% of the attempts in the training set, but some kickers have higher success rates than others. Instead of using the overall average for every attempt, we can use each kicker's training-set success rate. The code below adds that forecast as `phat_kicker`. Because we create a grouping for this calculation, we remove it with `ungroup()` when we are done.

```{r phat_kicker}
fg_train <- 
  fg_train %>% 
  group_by(Kicker) %>%
  mutate(phat_kicker = mean(Success)) %>%
  ungroup()
```

Distance is an important predictor of whether a field goal is made. We will use `cut()` to bin the attempts by distance, then compute the success rate across all kickers within each bin. This calculation works as intended because we removed the previous kicker grouping with `ungroup()`.

In our training set, the shortest attempt was 18 yards and the longest was 76 yards. We will begin with 10-yard intervals: $(10, 20]$, $(20, 30]$, ..., $(70, 80]$. We save the bin label in `Dist_10` and the corresponding prediction in `phat_dist_10`.

```{r phat_dist_10}
fg_train <-
  fg_train %>%
  mutate(Dist_10 = cut(Distance, breaks = seq(from = 10, to = 80, by = 10))) %>%
  group_by(Dist_10) %>%
  mutate(phat_dist_10 = mean(Success)) %>%
  ungroup()
```

We can now plot distance against `phat_dist_10`. Every attempt in the same bin receives the same prediction, so the plot forms a sequence of horizontal steps. We use alpha blending (see [Lecture 2](lecture2.html)) to make repeated points visible: locations with more overlapping attempts appear darker.

It certainly looks like our predictions make some intuitive sense: the estimated probability of making a field goal decreases as the distance increases. 

```{r plot phat_dist_10}
fg_plot <- ggplot(fg_train) +
  geom_point(aes(x = Distance, y = phat_dist_10), alpha = 0.2) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Distance (yards)", y = "Predicted probability")
fg_plot
```

We could instead use 5-yard or 2-yard intervals, saving the bin labels in `Dist_5` and `Dist_2`. As before, we add a prediction equal to the training-set success rate within each bin.

```{r dist_5 dist_2}
fg_train <- fg_train %>%
  mutate(Dist_5 = cut(Distance, breaks = seq(from = 10, to = 80, by = 5))) %>%
  group_by(Dist_5) %>%
  mutate(phat_dist_5 = mean(Success)) %>%
  ungroup()
fg_train <- fg_train %>%
  mutate(Dist_2 = cut(Distance, breaks = seq(from = 10, to = 80, by = 2))) %>%
  group_by(Dist_2) %>%
  mutate(phat_dist_2 = mean(Success)) %>%
  ungroup()
```

The faceted plot below makes it easier to compare the three bin widths. Each point represents the prediction for one observed distance.

```{r plot 3 bins}
fg_bin_predictions <- fg_train %>%
  select(Distance, phat_dist_10, phat_dist_5, phat_dist_2) %>%
  pivot_longer(
    cols = starts_with("phat_dist"),
    names_to = "model",
    values_to = "predicted_probability"
  ) %>%
  mutate(
    model = factor(
      model,
      levels = c("phat_dist_10", "phat_dist_5", "phat_dist_2"),
      labels = c("10-yard bins", "5-yard bins", "2-yard bins")
    )
  ) %>%
  distinct()

ggplot(fg_bin_predictions, aes(x = Distance, y = predicted_probability)) +
  geom_line(color = "dodgerblue4") +
  geom_point(color = "dodgerblue4", size = 1.5) +
  facet_wrap(vars(model), ncol = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Distance (yards)", y = "Predicted probability")
```

With 2-yard bins, the forecasts are no longer monotonically decreasing: although success generally becomes less likely as distance increases, a few adjacent bins move upward. Narrow bins use fewer attempts, so their averages are more sensitive to random variation. In [Lecture 8](lecture8.html), we will fit a smooth probability curve with logistic regression.


## Assessing Predictions

We now have several models of field-goal success: `phat_all`, `phat_kicker`, `phat_dist_10`, `phat_dist_5`, and `phat_dist_2`. Recall from [Lecture 5](lecture5.html) that we used RMSE to assess predictions of a continuous outcome. For a binary outcome, let $y_i$ be the outcome of observation $i$ and let $\hat{p}_i$ be its predicted probability of success.

The [Brier Score](https://en.wikipedia.org/wiki/Brier_score) is defined as
$$
BS = \frac{1}{n}\sum_{i = 1}^{n}{(y_{i} - \hat{p}_{i})^{2}}
$$
The Brier score is the mean squared error of the probability forecasts. Lower values indicate more accurate predictions. Using code similar to our RMSE calculation in [Lecture 5](lecture5.html), we can compute the training-set Brier score for each model.

```{r fg_train Brier}
reframe(fg_train,
  phat_all = mean((Success - phat_all)^2),
  phat_kicker = mean((Success - phat_kicker)^2),
  phat_dist_10 = mean((Success - phat_dist_10)^2),
  phat_dist_5 = mean((Success - phat_dist_5)^2),
  phat_dist_2 = mean((Success - phat_dist_2)^2)
)
```

Which model has the lowest training-set Brier score? Do you think this model may be overfitting?

Before assessing out-of-sample predictions, we will create tibbles that summarize each model's forecasts. For the kicker model, for example, we need only one row per kicker. The tibble `phat_kicker` therefore contains a kicker identifier and that kicker's training-set forecast:

```{r group_by kicker}
phat_kicker <- 
  fg_train %>%
  group_by(Kicker) %>%
  reframe(phat_kicker = mean(Success))
```

Create similar tibbles for `phat_dist_10`, `phat_dist_5`, and `phat_dist_2`.

```{r group_by dists}
phat_dist_10 <- 
  fg_train %>%
  group_by(Dist_10) %>%
  reframe(phat_dist_10 = mean(Success))
phat_dist_5 <- 
  fg_train %>%
  group_by(Dist_5) %>%
  reframe(phat_dist_5 = mean(Success))
phat_dist_2 <- 
  fg_train %>%
  group_by(Dist_2) %>%
  reframe(phat_dist_2 = mean(Success))
```

The file `nfl_fg_test.csv` contains held-out field goals from the same period. Because these attempts were not used to train our models, their predictions give us an estimate of *out-of-sample* performance. We first load the data into a tibble called `fg_test`.

```{r load fg_test, message = FALSE}
fg_test <- read_csv("data/nfl_fg_test.csv")
```

Using `mutate()` and `cut()`, add `Dist_10`, `Dist_5`, and `Dist_2` to `fg_test` using the same breaks as the training data.

```{r fg_test add bins}
fg_test <-
  fg_test %>%
  mutate(Dist_10 = cut(Distance, breaks = seq(from = 10, to = 80, by = 10)),
         Dist_5 = cut(Distance, breaks = seq(from = 10, to = 80, by = 5)),
         Dist_2 = cut(Distance, breaks = seq(from = 10, to = 80, by = 2)))
```

We are now ready to apply the forecasts learned from `fg_train` to `fg_test`. For `phat_all`, we use the overall success rate from `fg_train`—not the outcomes in `fg_test`.

```{r fg_test phat_all}
fg_test <-
  fg_test %>%
  mutate(phat_all = mean(fg_train$Success))
```

A join combines two tables using a *key*. Here, `Kicker` is the key: for each attempt in `fg_test`, the join finds the matching kicker in `phat_kicker` and appends that kicker's forecast. We use `left_join()` so every row of `fg_test` is retained.

```{r fg_test join}
fg_test <-
  fg_test %>%
  left_join(phat_kicker, by = "Kicker")
```

To verify that we have successfully performed this join, we can print out a few rows of `fg_test`:

```{r verify join}
select(fg_test, Kicker, Success, phat_all, phat_kicker)
```

Let's unpack the code. First, we overwrite `fg_test` with the result. We pipe `fg_test` to `left_join()`, pass `phat_kicker` as the table containing the additional data, and use `by = "Kicker"` to identify the matching key.

By mimicking the code above, we can add columns to `fg_test` for the remaining three predictive models: `phat_dist_10`, `phat_dist_5`, and `phat_dist_2`.

```{r fg_test add phat_dist}
fg_test <- 
  fg_test %>%
  left_join(phat_dist_10, by = "Dist_10") %>%
  left_join(phat_dist_5, by = "Dist_5") %>%
  left_join(phat_dist_2, by = "Dist_2")
```

Now, we can compute the out-of-sample Brier scores for each of our models. Which has the best out-of-sample performance?

```{r fg_test Brier, echo = FALSE}
reframe(fg_test,
  phat_all = mean((Success - phat_all)^2),
  phat_kicker = mean((Success - phat_kicker)^2),
  phat_dist_10 = mean((Success - phat_dist_10)^2),
  phat_dist_5 = mean((Success - phat_dist_5)^2),
  phat_dist_2 = mean((Success - phat_dist_2)^2)
)
```


## Looking Ahead

In [Lecture 8](lecture8.html), we will use *logistic regression* to replace these discrete bins with a smooth probability curve. To prepare, save the tibbles `fg_train` and `fg_test`:

```{r save train test}
save(fg_train, fg_test, file = "data/nfl_fg.RData")
```
