---
title: 'Lecture 8: Logistic Regression'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.align = "center")
library(tidyverse)
options(pillar.sigfig = 5, width = 100)
```

In [Lecture 7](lecture7.html), we estimated field-goal success by dividing distance into bins and averaging the outcomes within each bin. Narrow bins captured more detail but also followed more of the training sample's noise. Logistic regression replaces the steps with a smooth probability curve for a binary response.

## Why Logistic Regression?

Field-goal success is either 0 or 1. A linear regression can produce fitted values below 0 or above 1, which cannot be probabilities. Logistic regression fits a curve that stays between 0 and 1.

Load the [training and testing tibbles](data/nfl_fg.RData) saved at the end of [Lecture 7](lecture7.html).

```{r load nfl_fg, message=FALSE}
load("data/nfl_fg.RData")
```

## Logistic Regression with `glm()`

Fit a logistic regression with `glm()`. The formula still has the form `response ~ predictor`. The additional argument `family = binomial` tells R that the response is binary.

```{r logit distance}
logit_distance <- glm(
  Success ~ Distance,
  family = binomial,
  data = fg_train
)
```

To draw the curve, create a grid of distances and predict at every value. For logistic regression, `type = "response"` returns probabilities rather than the model's internal log-odds scale.

```{r distance prediction grid}
distance_grid <- tibble(
  Distance = seq(min(fg_train$Distance), max(fg_train$Distance), by = 1)
) %>%
  mutate(
    predicted_probability = predict(
      logit_distance,
      newdata = ., 
      type = "response"
    )
  )

ggplot(distance_grid, aes(x = Distance, y = predicted_probability)) +
  geom_line(linewidth = 1, color = "dodgerblue4") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Distance (yards)", y = "Predicted Probability of Success")
```

Short field goals have high fitted probabilities, and the probability decreases smoothly with distance.

## Comparing the Smooth and Binned Models

Add the logistic model's probabilities to both samples. The model is fitted only on `fg_train`; `newdata = fg_test` applies that fitted relationship to the held-out attempts.

```{r add distance predictions}
fg_train <- fg_train %>%
  mutate(
    phat_dist_logit = predict(
      logit_distance,
      newdata = fg_train,
      type = "response"
    )
  )

fg_test <- fg_test %>%
  mutate(
    phat_dist_logit = predict(
      logit_distance,
      newdata = fg_test,
      type = "response"
    )
  )
```

Calculate the Brier score for every model from Lecture 7 and for the new logistic model.

```{r distance model Brier comparison}
distance_train_scores <- fg_train %>%
  summarize(
    Overall = mean((Success - phat_all)^2),
    Kicker = mean((Success - phat_kicker)^2),
    `10-yard bins` = mean((Success - phat_dist_10)^2),
    `5-yard bins` = mean((Success - phat_dist_5)^2),
    `2-yard bins` = mean((Success - phat_dist_2)^2),
    `Logistic curve` = mean((Success - phat_dist_logit)^2)
  ) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "Brier") %>%
  mutate(Sample = "Training")

distance_test_scores <- fg_test %>%
  summarize(
    Overall = mean((Success - phat_all)^2),
    Kicker = mean((Success - phat_kicker)^2),
    `10-yard bins` = mean((Success - phat_dist_10)^2),
    `5-yard bins` = mean((Success - phat_dist_5)^2),
    `2-yard bins` = mean((Success - phat_dist_2)^2),
    `Logistic curve` = mean((Success - phat_dist_logit)^2)
  ) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "Brier") %>%
  mutate(Sample = "Testing")

bind_rows(distance_train_scores, distance_test_scores)
```

The narrow bins fit the training data more closely, but the logistic curve has the lowest testing Brier score. Its smoothness keeps it from following every small fluctuation in the training sample.

## Multiple Logistic Regression

A distance-only model assigns the same probability to every kicker attempting from the same distance. Add `Kicker` to fit separate curves for individual kickers.

```{r logit distance kicker}
logit_dist_kicker <- glm(
  Success ~ Distance + Kicker,
  family = binomial,
  data = fg_train
)
```

`Distance + Kicker` tells R to use both predictors. To visualize the model, create every combination of distance and three kickers, then predict for each row.

```{r kicker prediction grid}
dist_kicker_grid <- expand_grid(
  Distance = seq(min(fg_train$Distance), max(fg_train$Distance), by = 1),
  Kicker = c("Bailey", "Vinatieri", "Zuerlein")
) %>%
  mutate(
    predicted_probability = predict(
      logit_dist_kicker,
      newdata = ., 
      type = "response"
    )
  )

ggplot(
  dist_kicker_grid,
  aes(x = Distance, y = predicted_probability, color = Kicker)
) +
  geom_line(linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Distance (yards)", y = "Predicted Probability of Success")
```

The curves have the same shape but different heights. At the same distance, the model assigns Dan Bailey a higher fitted success probability than Greg Zuerlein or Adam Vinatieri.

## Does the Additional Predictor Help?

Add predictions from the multiple logistic regression and compare its Brier score with the distance-only model.

```{r multiple model predictions}
fg_train <- fg_train %>%
  mutate(
    phat_dist_kicker = predict(
      logit_dist_kicker,
      newdata = fg_train,
      type = "response"
    )
  )

fg_test <- fg_test %>%
  mutate(
    phat_dist_kicker = predict(
      logit_dist_kicker,
      newdata = fg_test,
      type = "response"
    )
  )
```

```{r final Brier comparison}
final_train_scores <- fg_train %>%
  summarize(
    `Distance only` = mean((Success - phat_dist_logit)^2),
    `Distance + kicker` = mean((Success - phat_dist_kicker)^2)
  ) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "Brier") %>%
  mutate(Sample = "Training")

final_test_scores <- fg_test %>%
  summarize(
    `Distance only` = mean((Success - phat_dist_logit)^2),
    `Distance + kicker` = mean((Success - phat_dist_kicker)^2)
  ) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "Brier") %>%
  mutate(Sample = "Testing")

final_scores <- bind_rows(final_train_scores, final_test_scores)
final_scores

ggplot(final_scores, aes(x = Model, y = Brier, fill = Sample)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "Brier Score", fill = NULL)
```

Adding the kicker lowers training error but slightly raises testing error. The additional predictor does not improve out-of-sample performance in this split.

Across Lectures 5--8, the central workflow has been: define a response, build candidate models on training data, visualize what each model is doing, and compare predictions on testing data. Complexity is useful only when it improves performance on observations the model did not fit.

Sports-data packages, plotting references, and project datasets are collected on the separate [Project Resources](project_resources.html) page.
