---
title: "Problem Set 4"
output:
  html_document:
    math_method: r-katex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
```

# Data visualization with nflfastR 

We will now apply our newly acquired plotting skills to make some visualizations working with data from nflfastR. If you have not installed `nflfastR` yet, run `install.packages("nflfastR")` once in the console. Then load in our usual libraries, as well as nflfastR.

```{r load in libraries, include = TRUE}
library(tidyverse)
library(nflfastR)
library(ggplot2)
```

1. Let's load in the data for the 2024 season. This dataset contains play-by-play data for all games in the 2024 NFL season.
```{r, include = TRUE}
pbp_2024 <- load_pbp(2024)
```

2. We can see that there are more than 100 columns. Let's tackle specific questions one at a time. First, look at quarterback completions by filtering for where `play_type == "pass"` and then selecting the following columns in a new table called `pbp_pass`.
```{r filtering, include = TRUE}
pbp_pass <- pbp_2024 %>% 
  filter(play_type == "pass") %>% 
  select(play_type, pass_location, yards_gained, air_yards, yards_after_catch, 
         passer_player_name, complete_pass, incomplete_pass, cpoe)

head(pbp_pass)
```

3. nflfastR unfortunately does not include a position field to identify quarterbacks. To work around this, group by player and keep players with more than 150 pass attempts to remove players without sufficient play time. Hint: pass attempts will be the sum of entries in `complete_pass` and `incomplete_pass`.
```{r filter for qbs}
pbp_pass <- pbp_pass %>% 
  group_by(passer_player_name) %>% 
  filter(sum(complete_pass, na.rm = TRUE) + sum(incomplete_pass, na.rm = TRUE) > 150) %>% 
  ungroup()
```

4. Let's take a preliminary look at how location might affect pass outcomes. First, use `!is.na()` with `filter()` to filter out pass attempts without a location label. Then create a bar plot to visualize the frequency of attempts in each pass location.

```{r bar plot}
pbp_pass <- pbp_pass %>% 
  filter(!is.na(pass_location))

ggplot(pbp_pass, aes(x = pass_location)) +
  geom_bar()
```

5. Let's look at the distribution of yards gained based on pass location with some box plots. Use `geom_boxplot()` to create a box plot of `yards_gained` by `pass_location`.
```{r}
ggplot(pbp_pass, aes(x = pass_location, y = yards_gained)) +
  geom_boxplot()
```

6. Completion percentage over expected is a metric that adjusts for the difficulty of a QB's throw based on throw timing, coverage, and location. Take a look at how `cpoe` differs based on pass location alone with `geom_violin()`.
```{r}
ggplot(pbp_pass, aes(x = pass_location, y = cpoe)) +
  geom_violin()
```

7. Let's now turn to aggregated pass statistics per quarterback. Group by `passer_player_name` and create a table called `pbp_grouped` with columns `passer_player_name`, `attempts`, `avg_yards_gained`, and `avg_cpoe` using `reframe()`.
```{r grouping}
pbp_grouped <- pbp_pass %>% 
  group_by(passer_player_name) %>% 
  reframe(
    attempts = sum(complete_pass, na.rm = TRUE) + sum(incomplete_pass, na.rm = TRUE),
    avg_yards_gained = mean(yards_gained, na.rm = TRUE),
    avg_cpoe = mean(cpoe, na.rm = TRUE)
  ) 

head(pbp_grouped)
```

8. First visualize how `attempts` and `avg_yards_gained` are related by using a scatterplot.
```{r scatter, include = TRUE}
ggplot(pbp_grouped, aes(x = attempts, y = avg_yards_gained)) +
  geom_point()
```

9. Next, add an abline to the plot using `geom_abline()` to see how linear the data may be. Use the `slope` and `intercept` arguments to set the slope to 0.005 and the intercept to 5. Set the linetype to dashed, and the color to black. This line was chosen by hand; in [Lecture 6](lecture6.html), you will learn how to fit regression models to the data mathematically to make a line of best fit.

```{r}
ggplot(pbp_grouped, aes(x = attempts, y = avg_yards_gained)) +
  geom_point(color = 'darkseagreen3', size = 3) +
  geom_abline(slope = 0.005, intercept = 5, linetype = "dashed", color = "black") +
  labs(
    title = "Quarterback Attempts vs Average Yards Gained",
    x = "Attempts",
    y = "Average Yards Gained"
  ) +
  theme_minimal()
```

10. Now, add a third variable. Set the color to equal `avg_cpoe` and use `scale_color_distiller()` to set the color scale.

```{r}
ggplot(pbp_grouped, aes(x = attempts, y = avg_yards_gained, color = avg_cpoe)) +
  geom_point(size = 3) +
  geom_abline(slope = 0.005, intercept = 5, linetype = "dashed", color = "black") +
  scale_color_distiller(palette = "Spectral") +
  labs(
    title = "Quarterback Attempts vs Average Yards Gained",
    x = "Attempts",
    y = "Average Yards Gained",
    color = "Average CPOE"
  ) +
  theme_minimal()
```

11. Finally, visualize the quarterbacks with the highest `avg_yards_gained`. Use `arrange()` and `slice_head()` to obtain the top 10 quarterbacks, then generate a bar plot with `geom_col()`. Make sure to either use 45 degree axis labels or flip the coordinates with `coord_flip()` so that the QB names are visible.

```{r top avg_yards_gained}
pbp_grouped %>% 
  arrange(desc(avg_yards_gained)) %>% 
  slice_head(n = 10) %>% 
  ggplot(aes(x = passer_player_name, y = avg_yards_gained)) +
  geom_col(fill = "darkseagreen3") +
  labs(
    title = "Top 10 Quarterbacks by Average Yards Gained",
    x = "Quarterback",
    y = "Average Yards Gained"
  ) +
  coord_flip() +
  theme_minimal()
```

12. Let's do one more small case study on explosive plays. First, start by filtering the data as below to get the appropriate columns.
```{r explosive filtering, include = TRUE}
pbp_explosive <- pbp_2024 %>%
  filter(play_type %in% c("pass", "run"),
         !is.na(yards_gained),
         !is.na(defteam),
         !is.na(down),
         !is.na(yardline_100)) %>%
  mutate(
    explosive = case_when(
      play_type == "run"  & yards_gained >= 10 ~ 1,
      play_type == "pass" & yards_gained >= 20 ~ 1,
      TRUE ~ 0
    )
  ) %>% 
  filter(explosive == 1) %>% 
  select(
    game_id,
    play_id,
    defteam,
    play_type,
    ydstogo,
    yardline_100,
    yards_gained,
    explosive
  )

head(pbp_explosive)
```

13. Visualize the distribution of explosive plays by defensive team. Use `geom_bar()` to create a bar plot of the number of explosive plays by `defteam`.
```{r}
ggplot(pbp_explosive, aes(x = defteam)) +
  geom_bar()
```

14. Take a closer look at the worst defenses against explosive plays. Group by defensive team and play type, then use `reframe()` to obtain the number of explosive plays in each category. Use `arrange()` and `slice_head()` to obtain the top 10 teams within each play type. Replot the bar graphs and facet by pass or run plays.

```{r}
pbp_explosive_slice <- pbp_explosive %>%
  group_by(defteam, play_type) %>%
  reframe(n_explosive = n()) %>%
  arrange(play_type, desc(n_explosive)) %>%
  group_by(play_type) %>%
  slice_head(n = 10)

ggplot(pbp_explosive_slice, aes(x = defteam, y = n_explosive)) +
  geom_col() +
  facet_wrap(~play_type)
```

15. Finally, visualize the relationship between explosive plays and yards to go with a heat map. First, run the following code below to obtain the needed columns for this plot.
```{r set up data table, include = TRUE}
pbp_explosive_rate <- pbp_2024 %>%
  filter(play_type %in% c("pass", "run"),
         !is.na(yards_gained),
         !is.na(defteam),
         !is.na(down),
         !is.na(yardline_100)) %>%
  mutate(
    explosive = case_when(
      play_type == "run"  & yards_gained >= 10 ~ 1,
      play_type == "pass" & yards_gained >= 20 ~ 1,
      TRUE ~ 0
    )
  ) %>% 
  select(
    game_id,
    play_id,
    defteam,
    play_type,
    ydstogo,
    yardline_100,
    yards_gained,
    explosive
  )
```

Then, group by `defteam` and `ydstogo` and obtain the explosive play *rate*, instead of number of explosive plays. Then use `geom_tile()` to create a heat map with `defteam` on the x-axis, `ydstogo` on the y-axis, and fill based on the explosive play rate.

```{r explosive rate}
pbp_explosive_rate <- pbp_explosive_rate %>%
  group_by(defteam, ydstogo) %>%
  reframe(
    n_explosive = sum(explosive),
    n_total = n(),
    explosive_rate = n_explosive / n_total
  ) %>%
  ungroup()
```

```{r heat map}
ggplot(pbp_explosive_rate, aes(x = defteam, y = ydstogo, fill = explosive_rate)) +
  geom_tile()
```
