---
title: "Problem Set 5"
output:
  html_document:
    math_method: r-katex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
```

```{r prepare data file, include=FALSE, eval=FALSE}
# Instructor-only recipe used to create data/nba_player_tpp.csv.
# The output is aggregated by player-season but is not filtered by season or attempts.
read_csv("data/nba_boxscore.csv", show_col_types = FALSE) %>%
  filter(Tm != "TOT") %>%
  group_by(Player, Season) %>%
  reframe(
    TPM = sum(TPM, na.rm = TRUE),
    TPA = sum(TPA, na.rm = TRUE)
  ) %>%
  mutate(TPP = if_else(TPA > 0, TPM / TPA, NA_real_)) %>%
  arrange(Season, Player) %>%
  write_csv("data/nba_player_tpp.csv")
```

# Predicting NBA Three-Point Percentage

In this problem set, we will predict an NBA player's 2016 three-point percentage using their 2015 three-point percentage. We will begin with the same prediction for every player and then divide the players into increasingly small groups based on their 2015 percentage.

The dataset contains one row for every NBA player-season from 1997 through 2016. It has **not** been filtered according to season or number of three-point attempts. Statistics for players who appeared on multiple teams in one season have already been combined so that each player has at most one row per season.

## Prepare the Data

1. Download [nba_player_tpp.csv](data/nba_player_tpp.csv), save it in your `data` folder, and run the following code to load it as a tbl called `shooting_raw`.

```{r load shooting, include=TRUE, message=FALSE}
shooting_raw <- read_csv("data/nba_player_tpp.csv")
```

The columns are:

* `Player`: player name
* `Season`: season
* `TPM`: three-pointers made
* `TPA`: three-pointers attempted
* `TPP`: three-point percentage

2. Use `filter()` to keep only the 2015 and 2016 seasons and player-seasons with at least 100 three-point attempts. Save the resulting tbl as `shooting_filtered`.

```{r filter seasons solution}
# solution
shooting_filtered <- shooting_raw %>%
  filter(Season %in% c(2015, 2016), TPA >= 100)
```

3. Some players met the attempt requirement in only one of the two seasons. Starting with `shooting_filtered`:

    * group the data by `Player`
    * use `filter()` and `n()` to keep players with two rows
    * ungroup the data
    * select `Player`, `Season`, `TPA`, and `TPP`
    * use `pivot_wider()` to put the 2015 and 2016 values in separate columns

Save the result as `shooting`. It should have one row per player and columns named `TPA_2015`, `TPA_2016`, `TPP_2015`, and `TPP_2016`.

```{r pivot shooting starter, include=TRUE, eval=FALSE}
shooting <- shooting_filtered %>%
  group_by(Player) %>%
  filter(...) %>%
  ungroup() %>%
  select(...) %>%
  pivot_wider(
    names_from = ...,
    values_from = ...
  )
```

```{r pivot shooting solution}
# solution
shooting <- shooting_filtered %>%
  group_by(Player) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  select(Player, Season, TPA, TPP) %>%
  pivot_wider(
    names_from = Season,
    values_from = c(TPA, TPP)
  )
```

4. Use `head()` to inspect `shooting`. Then make a scatterplot with `TPP_2015` on the horizontal axis and `TPP_2016` on the vertical axis. Use `alpha` to make overlapping points easier to see, and give the axes informative labels.

```{r initial scatter solution}
# solution
head(shooting)

ggplot(shooting) +
  geom_point(aes(x = TPP_2015, y = TPP_2016), alpha = 0.4) +
  labs(
    title = "NBA Three-Point Percentage in 2015 and 2016",
    x = "2015 Three-Point Percentage",
    y = "2016 Three-Point Percentage"
  ) +
  theme_minimal()
```

## Model 0: The Overall Mean

Suppose we had to predict a player's 2016 three-point percentage without using their 2015 percentage. A reasonable starting point would be the overall mean of `TPP_2016`.

5. Use `mutate()` and `mean()` to add a column called `yhat_0` containing this prediction for every player.

```{r overall mean solution}
# solution
shooting <- shooting %>%
  mutate(yhat_0 = mean(TPP_2016))
```

The prediction plots in this problem set use the same layered structure as the plots in Lecture 5: the first `geom_point()` layer shows the observed 2016 percentages, and the second `geom_point()` layer shows the model's predictions.

6. Create the first layered plot. Plot the observed `TPP_2016` values in transparent gray and the `yhat_0` predictions in red. Save the result as `plot_0`.

```{r overall mean plot solution}
# solution
plot_0 <- ggplot(shooting) +
  geom_point(aes(x = TPP_2015, y = TPP_2016), color = "gray40", alpha = 0.4) +
  geom_point(aes(x = TPP_2015, y = yhat_0), color = "lightcoral", size = 2) +
  labs(
    title = "Model 0: Overall Mean",
    x = "2015 Three-Point Percentage",
    y = "2016 Three-Point Percentage"
  ) +
  theme_minimal()

plot_0
```

7. Compute the root mean square error of `yhat_0`. Save the result as a one-row tbl called `rmse_0`.

$$RMSE = \sqrt{\text{mean}\left((\text{actual} - \text{prediction})^2\right)}$$

```{r overall mean rmse solution}
# solution
rmse_0 <- shooting %>%
  reframe(RMSE = sqrt(mean((TPP_2016 - yhat_0)^2)))

rmse_0
```

## Model 1: Two Groups

We can use 2015 three-point percentage by dividing the players into two groups: those below the overall 2015 average and those above it. Within each group, we will use the average `TPP_2016` as the prediction.

8. Use `cut()` to divide `TPP_2015` at its overall mean. Then group the data by the resulting intervals and add the prediction as a column called `yhat_1`. Remember to ungroup the data and remove the temporary `bins` column when you finish.

The endpoints 0.15 and 0.60 are slightly beyond the range of percentages in this dataset. The middle cut point should be `mean(shooting$TPP_2015)`.

```{r two groups starter, include=TRUE, eval=FALSE}
shooting <- shooting %>%
  mutate(bins = cut(
    TPP_2015,
    breaks = c(0.15, mean(shooting$TPP_2015), 0.60),
    include.lowest = TRUE
  )) %>%
  group_by(...) %>%
  mutate(yhat_1 = ...) %>%
  ungroup() %>%
  select(-bins)
```

```{r two groups solution}
# solution
shooting <- shooting %>%
  mutate(bins = cut(
    TPP_2015,
    breaks = c(0.15, mean(shooting$TPP_2015), 0.60),
    include.lowest = TRUE
  )) %>%
  group_by(bins) %>%
  mutate(yhat_1 = mean(TPP_2016)) %>%
  ungroup() %>%
  select(-bins)
```

9. Recreate the same layered plot, but replace `yhat_0` with `yhat_1`. Use blue for the prediction layer, change the title, and save the result as `plot_1`.

```{r two groups plot solution}
# solution
plot_1 <- ggplot(shooting) +
  geom_point(aes(x = TPP_2015, y = TPP_2016), color = "gray40", alpha = 0.4) +
  geom_point(aes(x = TPP_2015, y = yhat_1), color = "lightblue3", size = 2) +
  labs(
    title = "Model 1: Two Groups",
    x = "2015 Three-Point Percentage",
    y = "2016 Three-Point Percentage"
  ) +
  theme_minimal()

plot_1
```

## Models 2–4: Increasing the Number of Bins

We will now repeat the same process with increasingly narrow intervals. The `seq()` function will create evenly spaced cut points between 0.15 and 0.60.

10. Divide `TPP_2015` into intervals of width 0.05 and add the within-bin prediction as `yhat_2`. Then create `plot_2` using the same two layers as above, with green points for the predictions.

```{r width 005 starter, include=TRUE, eval=FALSE}
seq(from = 0.15, to = 0.60, by = 0.05)
```

```{r width 005 solution}
# solution
shooting <- shooting %>%
  mutate(bins = cut(
    TPP_2015,
    breaks = seq(from = 0.15, to = 0.60, by = 0.05),
    include.lowest = TRUE
  )) %>%
  group_by(bins) %>%
  mutate(yhat_2 = mean(TPP_2016)) %>%
  ungroup() %>%
  select(-bins)

plot_2 <- ggplot(shooting) +
  geom_point(aes(x = TPP_2015, y = TPP_2016), color = "gray40", alpha = 0.4) +
  geom_point(aes(x = TPP_2015, y = yhat_2), color = "darkseagreen3", size = 2) +
  labs(
    title = "Model 2: Bin Width 0.05",
    x = "2015 Three-Point Percentage",
    y = "2016 Three-Point Percentage"
  ) +
  theme_minimal()

plot_2
```

11. Repeat the process using intervals of width 0.025. Call the predictions `yhat_3`, use purple prediction points, and save the plot as `plot_3`.

```{r width 0025 solution}
# solution
shooting <- shooting %>%
  mutate(bins = cut(
    TPP_2015,
    breaks = seq(from = 0.15, to = 0.60, by = 0.025),
    include.lowest = TRUE
  )) %>%
  group_by(bins) %>%
  mutate(yhat_3 = mean(TPP_2016)) %>%
  ungroup() %>%
  select(-bins)

plot_3 <- ggplot(shooting) +
  geom_point(aes(x = TPP_2015, y = TPP_2016), color = "gray40", alpha = 0.4) +
  geom_point(aes(x = TPP_2015, y = yhat_3), color = "mediumpurple3", size = 2) +
  labs(
    title = "Model 3: Bin Width 0.025",
    x = "2015 Three-Point Percentage",
    y = "2016 Three-Point Percentage"
  ) +
  theme_minimal()

plot_3
```

12. Repeat the process one more time using intervals of width 0.01. Call the predictions `yhat_4`, use orange prediction points, and save the plot as `plot_4`.

```{r width 001 solution}
# solution
shooting <- shooting %>%
  mutate(bins = cut(
    TPP_2015,
    breaks = seq(from = 0.15, to = 0.60, by = 0.01),
    include.lowest = TRUE
  )) %>%
  group_by(bins) %>%
  mutate(yhat_4 = mean(TPP_2016)) %>%
  ungroup() %>%
  select(-bins)

plot_4 <- ggplot(shooting) +
  geom_point(aes(x = TPP_2015, y = TPP_2016), color = "gray40", alpha = 0.4) +
  geom_point(aes(x = TPP_2015, y = yhat_4), color = "darkorange", size = 2) +
  labs(
    title = "Model 4: Bin Width 0.01",
    x = "2015 Three-Point Percentage",
    y = "2016 Three-Point Percentage"
  ) +
  theme_minimal()

plot_4
```

## Compare the Predictions Visually

13. Run the code below to reshape the five prediction columns into a long tbl called `prediction_plot_data`. We will use this long tbl only for the faceted plot in this question.

```{r reshape predictions, include=TRUE, eval=FALSE}
prediction_plot_data <- shooting %>%
  select(Player, TPP_2015, TPP_2016, starts_with("yhat_")) %>%
  pivot_longer(
    cols = starts_with("yhat_"),
    names_to = "Model",
    values_to = "Prediction"
  )
```

```{r reshape predictions solution}
# solution
prediction_plot_data <- shooting %>%
  select(Player, TPP_2015, TPP_2016, starts_with("yhat_")) %>%
  pivot_longer(
    cols = starts_with("yhat_"),
    names_to = "Model",
    values_to = "Prediction"
  )
```

Use `prediction_plot_data` to make a faceted plot with one panel for each model. In every panel, use one point layer for the observed `TPP_2016` values in transparent gray and another point layer for the corresponding `Prediction` values in blue. Use `facet_wrap()` with `Model` as the faceting variable.

```{r faceted predictions solution}
# solution
ggplot(prediction_plot_data, aes(x = TPP_2015)) +
  geom_point(aes(y = TPP_2016), color = "gray40", alpha = 0.25) +
  geom_point(aes(y = Prediction), color = "dodgerblue3", size = 2) +
  facet_wrap(~Model, ncol = 2) +
  labs(
    title = "Predictions from Increasingly Fine Bins",
    x = "2015 Three-Point Percentage",
    y = "2016 Three-Point Percentage"
  ) +
  theme_minimal()
```

## Compare Predictive Error

14. Using `shooting`, compute the RMSE of all five prediction columns in a one-row tbl called `model_rmse`.

```{r all rmse solution}
# solution
model_rmse <- shooting %>%
  reframe(
    rmse_0 = sqrt(mean((TPP_2016 - yhat_0)^2)),
    rmse_1 = sqrt(mean((TPP_2016 - yhat_1)^2)),
    rmse_2 = sqrt(mean((TPP_2016 - yhat_2)^2)),
    rmse_3 = sqrt(mean((TPP_2016 - yhat_3)^2)),
    rmse_4 = sqrt(mean((TPP_2016 - yhat_4)^2))
  )

model_rmse
```

15. Reshape the one-row `model_rmse` tbl from Question 14 into a two-column tbl named `model_rmse_long`, with columns named `Model` and `RMSE`. Then make a point-and-line plot of RMSE across the five models. The model names should appear on the horizontal axis in order from `rmse_0` through `rmse_4`.

```{r rmse plot solution}
# solution
model_rmse_long <- model_rmse %>%
  pivot_longer(
    cols = everything(),
    names_to = "Model",
    values_to = "RMSE"
  )

ggplot(model_rmse_long, aes(x = Model, y = RMSE, group = 1)) +
  geom_line(color = "dodgerblue3", linewidth = 1) +
  geom_point(color = "dodgerblue3", size = 3) +
  labs(
    title = "RMSE on the Observed Data as the Bins Become Narrower",
    x = "Prediction Model",
    y = "RMSE"
  ) +
  theme_minimal()
```

These RMSE values are calculated using the same 163 players whose 2016 percentages were averaged to create the predictions. As the bins become narrower, the predictions can follow these observed outcomes increasingly closely. The RMSE plot demonstrates that fit to the observed data improves, but it does not tell us how the models would perform on a new season. In the next lecture, we will replace these abrupt jumps with predictions constrained to follow a straight line.
