---
title: "Problem Set 6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
```

# Regression with Triathlon Data

Now that we've seen linear regression in R applied to MLB batting averages, let's practice what we've learned in a context we haven't worked with before. We will look at past results of the Ironman Triathlon and investigate relationships between racing splits and demographic variables. First let us download the [Ironman Data](data/Ironman Data.csv) and save it to our "data" folder.

Let's read in the data and take a look at the variables.

```{r read Ironman data, include=TRUE}
data <- read_csv('data/Ironman Data.csv')
```

This is a large data set with hundreds of variables. Digging deeper, we can see that many of the variables seem to be repeated and many have data only in certain rows. In our case, this analysis will focus on the results from the 2017 Florida race, and we will use the swim, bike, run, and overall finishing times along with a few demographic variables.

The finishing times are stored as character strings, so we need to convert them to numeric values before using regression. Run the setup code below to create the cleaned dataset `results_2017`.

```{r clean Ironman data, include=TRUE, message=FALSE}
#install.packages("lubridate")
library(lubridate)
results_2017 <- data %>% 
  filter(`Source Table` == '2017 - Florida') %>% 
  select(Country, Gender, Division, Swim, Bike, Run, Overall) %>%
  drop_na() %>%
  mutate(Swim = period_to_seconds(ms(Swim))/60, 
         Bike = period_to_seconds(ms(Bike))/60, 
         Run = period_to_seconds(ms(Run))/60, 
         Overall = period_to_seconds(ms(Overall))/60)
```


## Simple Linear Regression

Now that we have our tidy dataset, let's start looking at the relationships between each leg of the triathlon.

1. Create scatterplots of swim times vs. bike times, bike times vs. run times, and swim times vs. run times. Save them as variables titled 'plot.swim_bike', etc.

```{r pt1 q1 solution}
# solution
plot.swim_bike<-ggplot(results_2017) +
  geom_point(aes(Swim, Bike))
plot.bike_run<-ggplot(results_2017) +
  geom_point(aes(Bike, Run))
plot.swim_run<-ggplot(results_2017) +
  geom_point(aes(Swim, Run))
```


2. Print the three plots you created above and compare the relationships visually.

```{r pt1 q2, include=TRUE}
plot.swim_bike
plot.bike_run
plot.swim_run
```

* Can you visually compare these associations by looking at the scatterplots? Which legs of the race seem to be more highly correlated?


3. Another way to display three continuous variables at once is by creating a scatterplot for two of the variables and then coloring the points by the values of the third variable. Try doing so with different combinations of our variables to create plots like the one below.

```{r pt1 q3 solution, include=TRUE, echo=FALSE}
# solution
ggplot(results_2017) +
  geom_point(aes(Bike, Swim, color = Run))
```

* From this one, we can see that the racers with faster swim and bike times tended to have faster run times; however, the specific relationship between Swim vs. Run and Bike vs. Run is less clear than in the individual scatter plots.


4. To numerically evaluate these relationships, compute the correlation between each of the variables (you should compute 3 different correlations). 

```{r pt1 q4 solution, include=TRUE, echo = FALSE}
# solution
cor(results_2017$Swim, results_2017$Bike)
cor(results_2017$Bike, results_2017$Run)
cor(results_2017$Swim, results_2017$Run)
```


5. Now, let's fit our linear model. Right now, we are doing simple linear regression, which involves one response variable and one predictor variable. So, let's start by running a regression with Swim as our predictor and Bike as our response.

```{r pt1 q5, include=TRUE, eval=FALSE}
lm.swim_bike <- lm( ... ) # fill this part in
```

```{r pt1 q5 solution}
# solution
lm.swim_bike <- lm(data = results_2017, formula = Bike ~ Swim)
```

* We obtain the following slope and intercept for this regression.

```{r  pt1 q5 2, include=TRUE}
lm.swim_bike$coefficients
```


6. We also want to add the regression line to our plot. Use `predict()` to generate predictions based on the model, and `mutate()` to add them to your data. Then, use `geom_line()` to add this predicted line to the plot. You may need to increase the size to increase the visibility of the line.
```{r pt1 q6, include=TRUE}
results_2017 <- results_2017 %>% 
  mutate(swim_bike_pred = predict(lm.swim_bike))

ggplot(results_2017) +
  geom_point(aes(Swim, Bike)) +
  geom_line(aes(x = Swim, y = swim_bike_pred), color = 'lightcoral', size = 1.5)
```


7. Repeat steps 5 and 6 for the two other combinations of the three race legs. Note that since we're only investigating associations between variables (as opposed to causal relations) it doesn't really matter which variable we choose to be our predictor and response; the resulting plots will be the same, except with flipped axes. If, instead, we were trying to determine causality, we would put the independent/predictor variable on the x-axis and the dependent/response variable on the y-axis.

```{r  pt1 q7 solution}
# solution
lm.bike_run <- lm(data = results_2017, formula = Run ~ Bike)
lm.bike_run$coefficients
results_2017 <- results_2017 %>% 
  mutate(bike_run_pred = predict(lm.bike_run))
ggplot(results_2017) +
  geom_point(aes(Bike, Run)) +
  geom_line( aes(x = Bike, y = bike_run_pred), color = 'lightcoral')

lm.run_swim <- lm(data = results_2017, formula = Swim ~ Run)
lm.run_swim$coefficients
results_2017 <- results_2017 %>% 
  mutate(run_swim_pred = predict(lm.run_swim))
ggplot(results_2017) +
  geom_point(aes(Run, Swim)) +
  geom_line(aes(x = Run, y = run_swim_pred), color = 'lightcoral')
```


After you finish this problem set, you should be comfortable fitting simple linear regression models, adding predicted lines to scatterplots, and comparing several linear relationships with correlation and RMSE.
