---
title: "Problem Set 6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.align = "center", include = FALSE)
library(tidyverse)
```

# Regression with Triathlon Data

The file [`ironman_florida_2017.csv`](data/ironman_florida_2017.csv) contains results from the 2017 Ironman Florida race. Each row is one athlete. Swim, bike, run, and overall times are measured in minutes.

## Explore the Relationship

1. Read the data and save it as `ironman`. Use `glimpse()` to inspect the variables.

```{r q1 prompt, include=TRUE, eval=FALSE}
ironman <- read_csv("data/ironman_florida_2017.csv")
glimpse(ironman)
```

```{r q1 solution}
ironman <- read_csv("data/ironman_florida_2017.csv", show_col_types = FALSE)
glimpse(ironman)
```

2. Plot bike time on the horizontal axis and run time on the vertical axis. Use transparent points so that dense parts of the plot remain visible. Give both axes clear labels with units.

```{r q2 solution}
ggplot(ironman, aes(x = Bike, y = Run)) +
  geom_point(alpha = 0.25) +
  labs(x = "Bike Time (minutes)", y = "Run Time (minutes)")
```

3. Calculate the correlation between bike and run times. Describe the direction and strength of the linear relationship using the scatterplot and the correlation together.

```{r q3 solution}
cor(ironman$Bike, ironman$Run)
```

## Fit and Interpret a Model

4. Fit a linear model with run time as the response and bike time as the predictor. Save it as `bike_model` and display its coefficients.

```{r q4 solution}
bike_model <- lm(Run ~ Bike, data = ironman)
bike_model$coefficients
```

5. Interpret the slope in minutes. The intercept corresponds to a bike time of zero minutes. Explain why that intercept is not meaningful for these data.

```{r q5 solution}
# The slope is the fitted difference in run time associated with a
# one-minute difference in bike time.

range(ironman$Bike)
# A bike time of zero is far outside the observed range.
```

6. Add the model's fitted values and residuals to `ironman`. Save them as `bike_fitted` and `bike_residual`.

```{r q6 solution}
ironman <- ironman %>%
  mutate(
    bike_fitted = predict(bike_model),
    bike_residual = Run - bike_fitted
  )
```

7. Recreate the scatterplot from Question 2 and add the fitted line with `geom_line()`. Remember to arrange the rows by bike time before drawing the line.

```{r q7 solution}
ggplot(arrange(ironman, Bike)) +
  geom_point(aes(x = Bike, y = Run), alpha = 0.25) +
  geom_line(
    aes(x = Bike, y = bike_fitted),
    color = "lightcoral",
    linewidth = 1
  ) +
  labs(x = "Bike Time (minutes)", y = "Run Time (minutes)")
```

8. Plot the residuals against the fitted values. Add a horizontal line at zero. Does the residual plot reveal a clear curve, a changing spread, or unusual observations?

```{r q8 solution}
ggplot(ironman, aes(x = bike_fitted, y = bike_residual)) +
  geom_hline(yintercept = 0, color = "lightcoral", linewidth = 1) +
  geom_point(alpha = 0.25) +
  labs(x = "Fitted Run Time (minutes)", y = "Residual (minutes)")
```

## Measure Fit and Predict New Rows

9. Calculate the model's $R^2$ and training RMSE. Interpret each value. RMSE is measured in minutes because run time is measured in minutes.

```{r q9 solution}
summary(bike_model)$r.squared

ironman %>%
  summarize(RMSE = sqrt(mean(bike_residual^2)))
```

10. Create a tibble called `new_bike_times` containing bike times of 300, 360, and 420 minutes. Use `predict()` with `newdata` to add each fitted run time.

```{r q10 solution}
new_bike_times <- tibble(Bike = c(300, 360, 420)) %>%
  mutate(predicted_Run = predict(bike_model, newdata = .))

new_bike_times
```

## Compare Two Predictors

11. Fit a second model with run time as the response and swim time as the predictor. Add its fitted values and residuals to `ironman`.

```{r q11 solution}
swim_model <- lm(Run ~ Swim, data = ironman)

ironman <- ironman %>%
  mutate(
    swim_fitted = predict(swim_model),
    swim_residual = Run - swim_fitted
  )
```

12. Create a table containing the $R^2$ and training RMSE for both models. Which predictor has the stronger linear relationship with run time? Support your answer with both measures.

```{r q12 solution}
model_comparison <- tibble(
  Model = c("Bike time", "Swim time"),
  R_squared = c(
    summary(bike_model)$r.squared,
    summary(swim_model)$r.squared
  ),
  RMSE = c(
    sqrt(mean(ironman$bike_residual^2)),
    sqrt(mean(ironman$swim_residual^2))
  )
)

model_comparison
```

13. Reshape `model_comparison` into a long table with one row for each model and measure. Plot the values in two side-by-side panels. Since $R^2$ and RMSE have different scales and units, give each panel its own vertical scale.

```{r q13 solution}
model_comparison %>%
  pivot_longer(
    cols = c(R_squared, RMSE),
    names_to = "Measure",
    values_to = "Value"
  ) %>%
  ggplot(aes(x = Model, y = Value, fill = Model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Measure, scales = "free_y") +
  labs(x = NULL, y = NULL)
```

After this problem set, you should be able to fit and interpret a simple linear regression, calculate fitted values and residuals, assess fit with $R^2$ and RMSE, use `newdata`, and compare two models.
