---
title: "Problem Set 7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.align = "center", include = FALSE)
library(tidyverse)
options(pillar.sigfig = 6, width = 100)
```

# Transformations and Multiple Regression

The files [`mlb_payrolls_train.csv`](data/mlb_payrolls_train.csv) and [`mlb_payrolls_test.csv`](data/mlb_payrolls_test.csv) contain observations for MLB teams and seasons with complete payroll and winning percentages for the current and previous seasons. The payroll source does not cover every team and season. The training file contains 1999--2018, and the testing file contains 2019 and 2021--2023. The shortened 2020 season is excluded as an outcome, but its winning percentages provide `previous_wp` for the 2021 rows.

The variables are:

- `team`: team name
- `season`: season of the outcome and payroll
- `winning_percentage`: wins divided by games played
- `relative_payroll`: team payroll divided by the MLB average in that season
- `previous_wp`: the team's winning percentage in the previous season

Relative payroll makes payroll comparable across seasons with different spending levels across MLB. Winning percentage from the previous season adds a measure of recent team performance that current payroll does not contain.

Fit all six candidate models with `mlb_train`. Load `mlb_test` only after all six models are fixed. Select the final model using testing RMSE, without refitting after viewing the scores.

## Candidate Transformations

1. Load `mlb_payrolls_train.csv` as `mlb_train`. Report the ranges of `season` and `relative_payroll`, then plot `relative_payroll` against `winning_percentage`.

```{r q1 prompt, include=TRUE, eval=FALSE}
mlb_train <- read_csv("data/mlb_payrolls_train.csv")
```

```{r q1 solution}
mlb_train <- read_csv(
  "data/mlb_payrolls_train.csv",
  show_col_types = FALSE
)

range(mlb_train$season)
range(mlb_train$relative_payroll)

ggplot(mlb_train, aes(x = relative_payroll, y = winning_percentage)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Payroll Relative to MLB Average",
    y = "Winning Percentage"
  )
```

2. Fit three models of `winning_percentage` using `relative_payroll`. Save them with the names below.

    - `lin_payroll_model`: `relative_payroll`
    - `log_payroll_model`: `log(relative_payroll)`
    - `quad_payroll_model`: `relative_payroll` and `relative_payroll^2`

    Use `I()` around the squared term.

```{r q2 starter, include=TRUE, eval=FALSE}
lin_payroll_model <- lm(winning_percentage ~ relative_payroll, data = mlb_train)
log_payroll_model <- lm(winning_percentage ~ log(relative_payroll), data = mlb_train)
quad_payroll_model <- lm(winning_percentage ~ ..., data = mlb_train)
```

```{r q2 solution}
lin_payroll_model <- lm(
  winning_percentage ~ relative_payroll,
  data = mlb_train
)

log_payroll_model <- lm(
  winning_percentage ~ log(relative_payroll),
  data = mlb_train
)

quad_payroll_model <- lm(
  winning_percentage ~ relative_payroll + I(relative_payroll^2),
  data = mlb_train
)
```

3. Create `payroll_grid`, a tibble containing 150 evenly spaced `relative_payroll` values from the observed minimum to the observed maximum. Use each model to add a column of fitted winning percentages. Reshape the three columns of fitted values into a long table called `payroll_grid_long`.

```{r q3 solution}
payroll_grid <- tibble(
  relative_payroll = seq(
    min(mlb_train$relative_payroll),
    max(mlb_train$relative_payroll),
    length.out = 150
  )
) %>%
  mutate(
    lin_payroll = predict(lin_payroll_model, newdata = .),
    log_payroll = predict(log_payroll_model, newdata = .),
    quad_payroll = predict(quad_payroll_model, newdata = .)
  )

payroll_grid_long <- payroll_grid %>%
  pivot_longer(
    cols = c(lin_payroll, log_payroll, quad_payroll),
    names_to = "model",
    values_to = "fitted_wp"
  ) %>%
  mutate(
    model = factor(
      model,
      levels = c("lin_payroll", "log_payroll", "quad_payroll"),
      labels = c("Lin. payroll", "Log payroll", "Quad. payroll")
    )
  )
```

4. Plot the observed training data and add all three curves from `payroll_grid_long`. Map `model` to color so that each curve has a different color and the plot includes a legend.

```{r q4 solution}
ggplot() +
  geom_point(
    data = mlb_train,
    mapping = aes(x = relative_payroll, y = winning_percentage),
    color = "gray40",
    alpha = 0.4
  ) +
  geom_line(
    data = payroll_grid_long,
    mapping = aes(
      x = relative_payroll,
      y = fitted_wp,
      color = model
    ),
    linewidth = 1
  ) +
  labs(
    x = "Payroll Relative to MLB Average",
    y = "Winning Percentage",
    color = NULL
  )
```

## Training Fit

5. Add fitted values from the three models to `mlb_train`, using the same fitted value column names as in `payroll_grid`.

```{r q5 solution}
mlb_train <- mlb_train %>%
  mutate(
    lin_payroll = predict(lin_payroll_model),
    log_payroll = predict(log_payroll_model),
    quad_payroll = predict(quad_payroll_model)
  )
```

6. Define `rmse()`, a function with `predicted` and `actual` arguments that returns RMSE. Use it to calculate the training RMSE for all three models and put the results in a long table called `payroll_train_scores`.

```{r q6 solution}
rmse <- function(predicted, actual) {
  sqrt(mean((actual - predicted)^2))
}

payroll_train_scores <- mlb_train %>%
  pivot_longer(
    cols = c(lin_payroll, log_payroll, quad_payroll),
    names_to = "model",
    values_to = "prediction"
  ) %>%
  reframe(
    rmse = rmse(prediction, winning_percentage),
    .by = model
  )

payroll_train_scores
```

7. Run the supplied code to add a column of random values, include it in the regression with linear payroll, and add the fitted values as `lin_payroll_random`. Use `rmse()` and `reframe()` to create `random_training_scores`, a long table comparing the training RMSE of `lin_payroll` and `lin_payroll_random`.

```{r q7 starter, include=TRUE, eval=FALSE}
set.seed(2026)

mlb_train <- mlb_train %>%
  mutate(random_value = rnorm(n()))

lin_payroll_random_model <- lm(
  winning_percentage ~ relative_payroll + random_value,
  data = mlb_train
)

mlb_train <- mlb_train %>%
  mutate(lin_payroll_random = predict(lin_payroll_random_model))
```

```{r q7 solution}
set.seed(2026)

mlb_train <- mlb_train %>%
  mutate(random_value = rnorm(n()))

lin_payroll_random_model <- lm(
  winning_percentage ~ relative_payroll + random_value,
  data = mlb_train
)

mlb_train <- mlb_train %>%
  mutate(lin_payroll_random = predict(lin_payroll_random_model))

random_training_scores <- mlb_train %>%
  reframe(
    lin_payroll = rmse(lin_payroll, winning_percentage),
    lin_payroll_random = rmse(lin_payroll_random, winning_percentage)
  ) %>%
  pivot_longer(
    everything(),
    names_to = "model",
    values_to = "rmse"
  )

random_training_scores
```

`random_value` contains no baseball information. Its model still has slightly lower training RMSE because `lm()` has another coefficient to fit. A lower training RMSE alone does not identify the better model. The random predictor model is not a candidate for testing.

## Adding Previous Season Winning Percentage

8. Calculate the residuals from `lin_payroll_model` as observed winning percentage minus `lin_payroll`. Plot the residuals against `previous_wp`, add a horizontal line at zero, and add a fitted straight line with `geom_smooth()`. The fitted line should slope upward.

```{r q8 solution}
training_residuals <- mlb_train %>%
  mutate(residual = winning_percentage - lin_payroll)

ggplot(training_residuals, aes(x = previous_wp, y = residual)) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_point(alpha = 0.35) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "lightcoral",
    linewidth = 1
  ) +
  labs(x = "Previous Season Winning Percentage", y = "Residual")
```

9. Fit three multiple regression models by adding the untransformed `previous_wp` predictor to each payroll specification. Save them as `lin_payroll_prev_wp_model`, `log_payroll_prev_wp_model`, and `quad_payroll_prev_wp_model`. Add their fitted values to `mlb_train` as `lin_payroll_prev_wp`, `log_payroll_prev_wp`, and `quad_payroll_prev_wp`.

```{r q9 solution}
lin_payroll_prev_wp_model <- lm(
  winning_percentage ~ relative_payroll + previous_wp,
  data = mlb_train
)

log_payroll_prev_wp_model <- lm(
  winning_percentage ~ log(relative_payroll) + previous_wp,
  data = mlb_train
)

quad_payroll_prev_wp_model <- lm(
  winning_percentage ~ relative_payroll + I(relative_payroll^2) +
    previous_wp,
  data = mlb_train
)

mlb_train <- mlb_train %>%
  mutate(
    lin_payroll_prev_wp = predict(lin_payroll_prev_wp_model),
    log_payroll_prev_wp = predict(log_payroll_prev_wp_model),
    quad_payroll_prev_wp = predict(quad_payroll_prev_wp_model)
  )
```

10. Display the coefficients of `log_payroll_prev_wp_model`. Calculate the fitted difference in current winning percentage associated with a 0.100 increase in `previous_wp`, holding relative payroll constant.

```{r q10 solution}
log_payroll_prev_wp_model$coefficients

coef(log_payroll_prev_wp_model)["previous_wp"] * 0.100
```

11. Calculate the training RMSE for the three multiple regression models. Combine these results with `payroll_train_scores` in a table called `training_scores`. Do not select a model or load the testing data.

```{r q11 solution}
multiple_train_scores <- mlb_train %>%
  pivot_longer(
    cols = c(
      lin_payroll_prev_wp,
      log_payroll_prev_wp,
      quad_payroll_prev_wp
    ),
    names_to = "model",
    values_to = "prediction"
  ) %>%
  reframe(
    rmse = rmse(prediction, winning_percentage),
    .by = model
  )

training_scores <- bind_rows(
  payroll_train_scores,
  multiple_train_scores
) %>%
  mutate(sample = "Training")

training_scores
```

## Testing on Later Seasons

The later seasons measure whether relationships fitted through 2018 carry forward in time.

12. Load `mlb_payrolls_test.csv` as `mlb_test`. Do not refit any model. Add predictions from all six fitted models.

```{r q12 solution}
mlb_test <- read_csv(
  "data/mlb_payrolls_test.csv",
  show_col_types = FALSE
) %>%
  mutate(
    lin_payroll = predict(lin_payroll_model, newdata = .),
    log_payroll = predict(log_payroll_model, newdata = .),
    quad_payroll = predict(quad_payroll_model, newdata = .),
    lin_payroll_prev_wp = predict(lin_payroll_prev_wp_model, newdata = .),
    log_payroll_prev_wp = predict(log_payroll_prev_wp_model, newdata = .),
    quad_payroll_prev_wp = predict(quad_payroll_prev_wp_model, newdata = .)
  )
```

13. Calculate testing RMSE for all six models and combine those scores with `training_scores`. Make a bar chart with the training and testing bars next to each other for each model, and label every bar with its RMSE. Create `best_payroll_model` containing the model using only payroll with the lowest testing RMSE and `best_overall_model` containing the model with the lowest testing RMSE overall.

```{r q13 solution}
testing_scores <- mlb_test %>%
  pivot_longer(
    cols = c(
      lin_payroll, log_payroll, quad_payroll,
      lin_payroll_prev_wp, log_payroll_prev_wp,
      quad_payroll_prev_wp
    ),
    names_to = "model",
    values_to = "prediction"
  ) %>%
  reframe(
    rmse = rmse(prediction, winning_percentage),
    .by = model
  ) %>%
  mutate(sample = "Testing")

model_comparison <- bind_rows(training_scores, testing_scores) %>%
  mutate(
    sample = factor(sample, levels = c("Training", "Testing")),
    model = factor(
      model,
      levels = c(
        "lin_payroll",
        "log_payroll",
        "quad_payroll",
        "lin_payroll_prev_wp",
        "log_payroll_prev_wp",
        "quad_payroll_prev_wp"
      ),
      labels = c(
        "Lin. payroll",
        "Log payroll",
        "Quad. payroll",
        "Lin. payroll + prev. win %",
        "Log payroll + prev. win %",
        "Quad. payroll + prev. win %"
      )
    )
  )

model_comparison

ggplot(
  model_comparison,
  aes(x = model, y = rmse, fill = sample)
) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = scales::number(rmse, accuracy = 0.0001)),
    position = position_dodge(width = 0.9),
    vjust = -0.4,
    size = 3
  ) +
  scale_x_discrete(
    labels = scales::label_wrap(18),
    guide = guide_axis(n.dodge = 2)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = NULL, y = "RMSE", fill = NULL)

best_payroll_model <- testing_scores %>%
  filter(!str_detect(model, "prev_wp")) %>%
  slice_min(rmse, n = 1)

best_overall_model <- testing_scores %>%
  slice_min(rmse, n = 1)

best_payroll_model
best_overall_model
```
