---
title: "Problem Set 7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.align = "center", include = FALSE)
library(tidyverse)
```

# Testing Regression Models

The files [`mlb_payrolls_train.csv`](data/mlb_payrolls_train.csv) and [`mlb_payrolls_test.csv`](data/mlb_payrolls_test.csv) contain team-season payroll and winning records. The rows were randomly divided into training and testing sets. We will compare two models of the relationship between team payroll and winning percentage.

## Training Data

1. Load `mlb_payrolls_train.csv` as `payroll_train`. Remove the shortened 2020 season. Use `glimpse()` and `range()` to inspect the variables and seasons.

```{r q1 solution}
payroll_train <- read_csv(
  "data/mlb_payrolls_train.csv",
  show_col_types = FALSE
) %>%
  filter(yearID != 2020)

glimpse(payroll_train)
range(payroll_train$yearID)
```

`payroll_norm` is a team's payroll divided by the average payroll in that season. A value of 1 is average, 2 is twice the average, and 0.5 is half the average. `log_payroll_norm` is the natural logarithm of that value.

2. Plot `payroll_norm` against winning percentage (`WP`). Label the New York Yankees and Oakland Athletics separately from all other teams, and color the points by that label.

```{r q2 solution}
payroll_train <- payroll_train %>%
  mutate(
    team_label = case_when(
      name == "New York Yankees" ~ "New York Yankees",
      name == "Oakland Athletics" ~ "Oakland Athletics",
      TRUE ~ "Other"
    )
  )

ggplot(payroll_train, aes(x = payroll_norm, y = WP, color = team_label)) +
  geom_point(alpha = 0.55) +
  labs(
    x = "Payroll Relative to League Average",
    y = "Winning Percentage",
    color = "Team"
  )
```

3. Fit two models using only `payroll_train`:

- `raw_model`: winning percentage as a function of `payroll_norm`
- `log_model`: winning percentage as a function of `log_payroll_norm`

Add both models' fitted values to `payroll_train` as `raw_fitted` and `log_fitted`.

```{r q3 solution}
raw_model <- lm(WP ~ payroll_norm, data = payroll_train)
log_model <- lm(WP ~ log_payroll_norm, data = payroll_train)

payroll_train <- payroll_train %>%
  mutate(
    raw_fitted = predict(raw_model),
    log_fitted = predict(log_model)
  )
```

4. Reshape the two fitted-value columns into a long table. Plot the observed data and the two fitted lines in separate panels. Arrange the rows by `payroll_norm` before drawing the lines.

```{r q4 solution}
payroll_fitted_long <- payroll_train %>%
  select(payroll_norm, WP, raw_fitted, log_fitted) %>%
  pivot_longer(
    cols = c(raw_fitted, log_fitted),
    names_to = "Model",
    values_to = "Fitted"
  ) %>%
  mutate(
    Model = recode(
      Model,
      raw_fitted = "Raw payroll",
      log_fitted = "Log payroll"
    )
  )

ggplot(arrange(payroll_fitted_long, payroll_norm)) +
  geom_point(aes(x = payroll_norm, y = WP), alpha = 0.35) +
  geom_line(
    aes(x = payroll_norm, y = Fitted),
    color = "lightcoral",
    linewidth = 1
  ) +
  facet_wrap(~ Model) +
  labs(x = "Payroll Relative to League Average", y = "Winning Percentage")
```

5. Calculate a residual for each model. Reshape the residuals into a long table and make a faceted residual plot with a horizontal line at zero. Do either of the plots show a clear pattern?

```{r q5 solution}
payroll_residuals_long <- payroll_train %>%
  transmute(
    payroll_norm,
    `Raw payroll` = WP - raw_fitted,
    `Log payroll` = WP - log_fitted
  ) %>%
  pivot_longer(
    cols = c(`Raw payroll`, `Log payroll`),
    names_to = "Model",
    values_to = "Residual"
  )

ggplot(payroll_residuals_long, aes(x = payroll_norm, y = Residual)) +
  geom_hline(yintercept = 0, color = "lightcoral", linewidth = 1) +
  geom_point(alpha = 0.4) +
  facet_wrap(~ Model) +
  labs(x = "Payroll Relative to League Average", y = "Residual")
```

## Training and Testing Error

6. Add a baseline model that predicts the mean winning percentage from `payroll_train` for every team-season. Calculate the training RMSE for the baseline, raw-payroll, and log-payroll models.

```{r q6 solution}
training_scores <- payroll_train %>%
  summarize(
    Baseline = sqrt(mean((WP - mean(WP))^2)),
    `Raw payroll` = sqrt(mean((WP - raw_fitted)^2)),
    `Log payroll` = sqrt(mean((WP - log_fitted)^2))
  ) %>%
  pivot_longer(
    everything(),
    names_to = "Model",
    values_to = "RMSE"
  ) %>%
  mutate(Sample = "Training")

training_scores
```

7. Load `mlb_payrolls_test.csv` as `payroll_test` and remove 2020. Do not refit either model. Use the models fitted in Question 3 to add `raw_prediction` and `log_prediction`. Add the training-set mean as `baseline_prediction`.

```{r q7 solution}
payroll_test <- read_csv(
  "data/mlb_payrolls_test.csv",
  show_col_types = FALSE
) %>%
  filter(yearID != 2020) %>%
  mutate(
    baseline_prediction = mean(payroll_train$WP),
    raw_prediction = predict(raw_model, newdata = .),
    log_prediction = predict(log_model, newdata = .)
  )
```

8. Calculate the testing RMSE for all three models. Combine the training and testing scores in one table, then plot them. Which model has the lowest testing RMSE? How large is its advantage over the other payroll model?

```{r q8 solution}
testing_scores <- payroll_test %>%
  summarize(
    Baseline = sqrt(mean((WP - baseline_prediction)^2)),
    `Raw payroll` = sqrt(mean((WP - raw_prediction)^2)),
    `Log payroll` = sqrt(mean((WP - log_prediction)^2))
  ) %>%
  pivot_longer(
    everything(),
    names_to = "Model",
    values_to = "RMSE"
  ) %>%
  mutate(Sample = "Testing")

rmse_comparison <- bind_rows(training_scores, testing_scores)
rmse_comparison

ggplot(rmse_comparison, aes(x = Model, y = RMSE, fill = Sample)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "RMSE", fill = NULL)
```

The two payroll models should have very similar testing RMSEs. A small numerical difference is not strong evidence that one functional form will always predict better.

## Challenge

Use the payroll model with the lowest testing RMSE. For each test-set team-season, convert the prediction error in winning percentage to wins over 162 games. Then calculate each team's average wins above or below its prediction and make a horizontal bar plot.

```{r challenge solution}
# The log-payroll model has the slightly lower testing RMSE in this split.
payroll_test_summary <- payroll_test %>%
  mutate(wins_vs_prediction = (WP - log_prediction) * 162) %>%
  group_by(name) %>%
  summarize(mean_wins_vs_prediction = mean(wins_vs_prediction))

ggplot(
  payroll_test_summary,
  aes(x = reorder(name, mean_wins_vs_prediction), y = mean_wins_vs_prediction)
) +
  geom_col(fill = "lightblue4", alpha = 0.7) +
  coord_flip() +
  labs(
    x = "Team",
    y = "Average Wins Above or Below Prediction"
  )
```
