---
title: "Problem Set 7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
library(ggplot2)
```


# Testing Regression Models

We will now continue modeling data with regression, but we can now test and compare different models more efficiently. We will use this dataset on mlb payrolls: [mlb_payrolls_train.csv](data/mlb_payrolls_train.csv) to analyze general manager performance based on team payrolls and winning percentages.

1. First, load in the data to R. Let's see the range of seasons that are included in this data 
```{r, include = TRUE}
payroll <- read_csv("data/mlb_payrolls_train.csv")

unique(payroll$yearID)
```
1. We can see that the data ranges from 1998 to 2023. Since the 2020 season was shortened due to Covid, filter it out. Then use `mutate()` to create a column named `WP` based on wins (W) and games (G). 

```{r column names}
head(payroll)
```


```{r filter and mutate}
payroll <- payroll %>% 
  filter(yearID != 2020) %>% 
  mutate(WP = W/G)
```


2. Then, explore the relationship between payrolls and winning percentage by generating a scatterplot. Use `mutate()` and `case_when()` to create a column that labels the Oakland A's and New York Yankees separately from the other teams, then color the points by this new label.
```{r plot}
payroll <- payroll %>%
  mutate(team_label = case_when(
    name == "New York Yankees" ~ "New York Yankees",
    name == "Oakland Athletics" ~ "Oakland Athletics",
    TRUE ~ "Other"
  ))

ggplot(payroll, aes(x = payroll_norm, y = WP, color = team_label)) +
  geom_point(size = 3, alpha = 0.5) +
  labs(
    title = "MLB Payrolls vs Winning Percentage",
    x = "Normalized Payroll",
    y = "Winning Percentage",
    color = "Team"
  ) +
  theme_minimal()

```

3. Now, fit a regression line to the data as a function of `payroll_norm`, then add it to the plot. 
```{r}
payroll_model <- lm(WP ~ payroll_norm, data = payroll)

payroll <- payroll %>%
  mutate(predicted_wp = predict(payroll_model))

# Plot with predicted line
ggplot(payroll, aes(x = payroll_norm, y = WP)) +
  geom_point() +
  geom_line(aes(y = predicted_wp), color = "lightcoral", linewidth = 1) +
  labs(
    title = "MLB Payrolls vs Winning Percentage (with Regression Line)",
    x = "Normalized Payroll",
    y = "Winning Percentage"
  ) +
  theme_minimal()
```

4. Next, try fitting a model based on `log_payroll_norm` and adding that to the plot. 
```{r log model}
log_model <- lm(WP ~ log_payroll_norm, data = payroll)
payroll <- payroll %>%
  mutate(log_predicted_wp = predict(log_model))
ggplot(payroll, aes(x = payroll_norm, y = WP)) +
  geom_point() +
  geom_line(aes(y = log_predicted_wp), color = "lightcoral", linewidth = 1) +
  labs(
    title = "MLB Payrolls vs Winning Percentage (with Regression Line)",
    x = "Normalized Payroll",
    y = "Winning Percentage"
  ) +
  theme_minimal()
```
Just by looking at the plots from 3 and 4, which seems to fit the data better? 

5. In Lecture 6, we defined a residual as the observed outcome minus the predicted outcome. Use `mutate()` to add a residual column for each model to `payroll`. Then generate a scatter plot of the residuals from both models.
```{r}
payroll <- payroll %>%
  mutate(
    residuals = WP - predicted_wp,
    log_residuals = WP - log_predicted_wp
  )

ggplot(payroll, aes(x = payroll_norm)) +
  geom_point(aes(y = residuals), color = "lightblue4") +
  geom_point(aes(y = log_residuals), color = "lightcoral") +
  labs(
    title = "Residuals of Payroll Models",
    x = "Normalized Payroll",
    y = "Residuals"
  ) +
  theme_minimal() 
```

An indication of a poor model fit is a non-uniform residual plot. This would indicate that perhaps a linear model is being used when the underlying trend is instead logarithmic, or polynomial. Do you observe this? 

6. Finally, we can directly compare our models with out-of-sample performance. Load in [mlb_payrolls_test.csv](data/mlb_payrolls_test.csv) and call it `payrolls_test`. Do not refit your model but instead use `predict` with the models you have already created to generate predictions for the test data set called `test_pred` and `test_log_pred`, similar to its usage in Lecture 6. Then, use the line below to compute the RMSE for each model. 
```{r}
payrolls_test <- read_csv("data/mlb_payrolls_test.csv")
payrolls_test <- payrolls_test %>%
  mutate(
    test_pred = predict(payroll_model, newdata = payrolls_test),
    test_log_pred = predict(log_model, newdata = payrolls_test)
  )
```

```{r RMSE, include = TRUE}
rmse <- sqrt(mean((payrolls_test$WP - payrolls_test$test_pred)^2))
rmse_log <- sqrt(mean((payrolls_test$WP - payrolls_test$test_log_pred)^2))
```

```{r values}
rmse
rmse_log
```

Based on the RMSE values, which model is better? 

*Challenge*: Find the difference between the actual WP values and predictions of the *log model* in the test data and add them to `payrolls_test`. Group by team and find the mean difference across all years. Multiply their percentages by 162 to obtain number of wins. Make a bar plot showing teams' mean performance compared to predictions.
```{r}
payrolls_test_summary <- payrolls_test %>%
  mutate(
    diff_wp = WP - test_log_pred  
  ) %>%
  group_by(name) %>%
  reframe(
    mean_diff_wp = mean(diff_wp, na.rm = TRUE)
  ) %>%
  mutate(
    wins_from_prediction = mean_diff_wp * 162  # convert WP difference to win difference
  )

# Plot
ggplot(payrolls_test_summary, aes(x = name, y = wins_from_prediction)) +
  geom_col(fill = "lightblue4", alpha = 0.6) +
  labs(
    title = "Average Wins Above/Below Prediction by Team",
    x = "Team",
    y = "Wins vs Predicted (Payroll Model)"
  ) +
  theme_minimal() +
  coord_flip()
```
