---
title: "Problem Set 8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
library(ggplot2)
```


# Logistic Regression

We will now continue modeling data with logistic regression using [nfl_wp.csv](data/nfl_wp.csv) to analyze win percentages as a function of multiple variables. The dataset includes `game_id`; `season`; `label_win`, which is a binary (0 or 1) indicator on whether the possessing team won the game; `score_differential`, which is the current score differential between the team in possession and the team on defense; `posteam_spread`, which is the +- spread for the possessing team prior to the game start, `yardline_100`; `down`; and `ydstogo`. 

1. First, load in the training and testing data.
```{r split data, include = TRUE}
nfl_train <- read_csv("data/nfl_wp_train.csv")
nfl_test <- read_csv("data/nfl_wp_test.csv")
```


2. Next, do some exploratory data analysis (EDA) to understand the data. First, plot a histogram of `score_differential` to show the distribution of the variable. Then, make a separate data table grouped by game and select the first instance of `posteam_spread` so that we have only one per game. Plot the distribution of this variable.

```{r}

nfl_wp_grouped <- nfl_train %>%
  group_by(game_id) %>%
  reframe(
    spread = first(posteam_spread)
  ) 

ggplot(nfl_wp_grouped, aes(x = spread)) +
  geom_histogram(binwidth = 2, fill = "lightblue4") +
  labs(title = "Distribution of Team Spread",
       x = "Spread",
       y = "Count") +
  theme_minimal()

ggplot(nfl_train, aes(x = score_differential)) +
  geom_histogram(binwidth = 1, fill = "lightcoral") +
  labs(title = "Distribution of Score Differential",
       x = "Spread",
       y = "Count") +
  theme_minimal()
```

3. Then fit a logistic model called `diff_model` using `glm()` to predict `label_win` based on `score_differential`. Call the model after you fit it to see the coefficients and then add its predictions to the training dataset.
```{r model and predictions.}
diff_model = glm(label_win ~ score_differential, family = binomial, data = nfl_train)
diff_model 
nfl_train <- nfl_train %>% 
  mutate(diff_predicted_win = predict(diff_model, type = "response"))
nfl_train
```

4. Repeat step 3 but predicting with `posteam_spread` instead. 
```{r}
spread_model = glm(label_win ~ posteam_spread, family = binomial, data = nfl_train)
spread_model 
nfl_train <- nfl_train %>% 
  mutate(spread_predicted_win = predict(spread_model, type = "response"))
nfl_train
```

5. Now, plot the predicted probabilities of winning based on the two models you created in steps 3 and 4. Put the predictor variable on the x-axis and the predicted win probability on the y axis. Create a label column with `case_when()` so that the points are colored by whether the team eventually won or lost. Which variable seems to predict better? 
```{r}
nfl_train <- nfl_train %>%
  mutate(win_label = case_when(
    label_win == 1 ~ "Win",
    TRUE ~ "Loss"
  ))

ggplot(nfl_train, aes(x = score_differential, y = diff_predicted_win, color = win_label)) +
  geom_point(alpha = 0.5) +
  labs(title = "Predicted Win Probability vs Score Differential",
       x = "Score Differential",
       y = "Predicted Win Probability",
       color = "Actual Win") +
  theme_minimal()

ggplot(nfl_train, aes(x = posteam_spread, y = spread_predicted_win, color = win_label)) +
  geom_point(alpha = 0.5) +
  labs(title = "Predicted Win Probability vs Spread",
       x = "Spread",
       y = "Predicted Win Probability",
       color = "Actual Win") +
  theme_minimal()
```

6. Finally, let's make a model to account for both variables. Name your model `spread_diff_model` and follow the previous steps to obtain predictions and plot the data as a function of score differential (since this variable is more continuous than spread as we saw with our histogram). How does this model compare to the previous two?
```{r}
spread_diff_model = glm(label_win ~ score_differential + posteam_spread, family = binomial, data = nfl_train)
nfl_train <- nfl_train %>% 
  mutate(spread_diff_predicted_win = predict(spread_diff_model, type = "response"))
ggplot(nfl_train, aes(x = score_differential, y = spread_diff_predicted_win, color = win_label)) +
  geom_point(alpha = 0.5) +
  labs(title = "Predicted Win Probability vs Score Differential (with Spread)",
       x = "Score Differential",
       y = "Predicted Win Probability",
       color = "Actual Win") +
  theme_minimal()
```

7. Let's test all of our models. Use `predict()` on the test data with the previously created models, then add them to the test dataset. Then, run the following code below to convert from win probability to a predicted win or loss and then calculate the accuracy. Print your results. Which model ended up being the best?
```{r predictions}
nfl_test <- nfl_test %>%
  mutate(
    diff_pred_prob = predict(diff_model, newdata = nfl_test, type = "response"),
    spread_pred_prob = predict(spread_model, newdata = nfl_test, type = "response"),
    spread_diff_pred_prob = predict(spread_diff_model, newdata = nfl_test, type = "response")
  )

```

```{r accuracy, include = TRUE}
nfl_test <- nfl_test %>% 
  mutate(
    diff_pred_label = case_when(diff_pred_prob >= 0.5 ~ 1, TRUE ~ 0),
    spread_pred_label = case_when(spread_pred_prob >= 0.5 ~ 1, TRUE ~ 0),
    spread_diff_pred_label = case_when(spread_diff_pred_prob >= 0.5 ~ 1, TRUE ~ 0)
  )

acc_diff  <- mean(nfl_test$label_win == nfl_test$diff_pred_label)
acc_spread <- mean(nfl_test$label_win == nfl_test$spread_pred_label)
acc_both  <- mean(nfl_test$label_win == nfl_test$spread_diff_pred_label)

```

```{r results}

tibble(
  Model = c("Score Differential", "Spread", "Score Differential + Spread"),
  Accuracy = c(acc_diff, acc_spread, acc_both)
)
```

*Challenge: Can you make an even better model with the other variables in the dataset or with numerical transformations to variables? Additionally, feel free to explore other metrics for out-of-sample performance, such as log-loss or AUC!*
