---
title: "Problem Set 8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.align = "center", include = FALSE)
library(tidyverse)
```

# Logistic Regression and Win Probability

The data contain one row for each NFL play. `label_win` indicates whether the team in possession eventually won the game. `score_differential` is that team's current lead or deficit, and `posteam_spread` is its pregame point spread. Negative spreads indicate a pregame favorite.

The training file contains the 2016--2020 seasons. The testing file contains the 2021 season. Games do not appear in both samples.

## Explore the Data

1. Load [`nfl_wp_train.csv`](data/nfl_wp_train.csv) as `nfl_train` and [`nfl_wp_test.csv`](data/nfl_wp_test.csv) as `nfl_test`. Count the rows and distinct games in each sample. Confirm that their game IDs do not overlap.

```{r q1 prompt, include=TRUE, eval=FALSE}
nfl_train <- read_csv("data/nfl_wp_train.csv")
nfl_test <- read_csv("data/nfl_wp_test.csv")
```

```{r q1 solution}
nfl_train <- read_csv("data/nfl_wp_train.csv", show_col_types = FALSE)
nfl_test <- read_csv("data/nfl_wp_test.csv", show_col_types = FALSE)

nfl_train %>%
  summarize(plays = n(), games = n_distinct(game_id))

nfl_test %>%
  summarize(plays = n(), games = n_distinct(game_id))

intersect(unique(nfl_train$game_id), unique(nfl_test$game_id))
```

2. Plot the training distribution of `score_differential`. Then use `distinct(game_id, posteam_spread)` to keep one row per game and team spread before plotting the distribution of `posteam_spread`. Give both plots accurate axis labels.

```{r q2 solution}
ggplot(nfl_train, aes(x = score_differential)) +
  geom_histogram(binwidth = 2, fill = "lightcoral") +
  labs(x = "Score Differential", y = "Plays")

game_team_spreads <- nfl_train %>%
  distinct(game_id, posteam_spread)

ggplot(game_team_spreads, aes(x = posteam_spread)) +
  geom_histogram(binwidth = 1, fill = "lightblue4") +
  labs(x = "Pregame Team Spread", y = "Team-Games")
```

## Fit and Visualize Logistic Models

3. Fit `diff_model`, a logistic regression of `label_win` on `score_differential`. Create a grid covering the observed score differentials, add predicted probabilities with `type = "response"`, and plot the fitted curve.

```{r q3 solution}
diff_model <- glm(
  label_win ~ score_differential,
  family = binomial,
  data = nfl_train
)

score_grid <- tibble(
  score_differential = seq(
    min(nfl_train$score_differential),
    max(nfl_train$score_differential),
    by = 1
  )
) %>%
  mutate(
    predicted_win_probability = predict(
      diff_model,
      newdata = ., 
      type = "response"
    )
  )

ggplot(score_grid, aes(x = score_differential, y = predicted_win_probability)) +
  geom_line(color = "dodgerblue4", linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Score Differential", y = "Predicted Win Probability")
```

4. Fit `spread_model`, a logistic regression of `label_win` on `posteam_spread`. Create a prediction grid and plot its fitted probability curve. How does being a larger pregame favorite change the fitted win probability?

```{r q4 solution}
spread_model <- glm(
  label_win ~ posteam_spread,
  family = binomial,
  data = nfl_train
)

spread_grid <- tibble(
  posteam_spread = seq(
    min(nfl_train$posteam_spread),
    max(nfl_train$posteam_spread),
    by = 0.5
  )
) %>%
  mutate(
    predicted_win_probability = predict(
      spread_model,
      newdata = ., 
      type = "response"
    )
  )

ggplot(spread_grid, aes(x = posteam_spread, y = predicted_win_probability)) +
  geom_line(color = "dodgerblue4", linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Pregame Team Spread", y = "Predicted Win Probability")
```

5. Fit `combined_model` using both `score_differential` and `posteam_spread`. Create a grid containing every score differential and three pregame spreads: -7, 0, and 7. Predict on the grid and plot three probability curves. Explain how the pregame spread changes win probability for teams with the same current score differential.

```{r q5 solution}
combined_model <- glm(
  label_win ~ score_differential + posteam_spread,
  family = binomial,
  data = nfl_train
)

combined_grid <- expand_grid(
  score_differential = seq(
    min(nfl_train$score_differential),
    max(nfl_train$score_differential),
    by = 1
  ),
  posteam_spread = c(-7, 0, 7)
) %>%
  mutate(
    predicted_win_probability = predict(
      combined_model,
      newdata = ., 
      type = "response"
    ),
    spread_label = factor(posteam_spread)
  )

ggplot(
  combined_grid,
  aes(
    x = score_differential,
    y = predicted_win_probability,
    color = spread_label
  )
) +
  geom_line(linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Score Differential",
    y = "Predicted Win Probability",
    color = "Pregame Spread"
  )
```

## Compare Training and Testing Performance

6. Use all three fitted models to add predicted probabilities to both `nfl_train` and `nfl_test`. Do not refit the models on the testing data.

```{r q6 solution}
nfl_train <- nfl_train %>%
  mutate(
    `Score differential` = predict(diff_model, newdata = ., type = "response"),
    `Pregame spread` = predict(spread_model, newdata = ., type = "response"),
    Combined = predict(combined_model, newdata = ., type = "response")
  )

nfl_test <- nfl_test %>%
  mutate(
    `Score differential` = predict(diff_model, newdata = ., type = "response"),
    `Pregame spread` = predict(spread_model, newdata = ., type = "response"),
    Combined = predict(combined_model, newdata = ., type = "response")
  )
```

7. Calculate the training and testing Brier scores for all three models. Combine them in one table and plot them. Which model has the lowest testing Brier score? Does the ranking match the training results?

```{r q7 solution}
training_scores <- nfl_train %>%
  summarize(
    across(
      c(`Score differential`, `Pregame spread`, Combined),
      ~ mean((label_win - .x)^2)
    )
  ) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "Brier") %>%
  mutate(Sample = "Training")

testing_scores <- nfl_test %>%
  summarize(
    across(
      c(`Score differential`, `Pregame spread`, Combined),
      ~ mean((label_win - .x)^2)
    )
  ) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "Brier") %>%
  mutate(Sample = "Testing")

brier_comparison <- bind_rows(training_scores, testing_scores)
brier_comparison

ggplot(brier_comparison, aes(x = Model, y = Brier, fill = Sample)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "Brier Score", fill = NULL)
```

## Challenge

Fit another logistic regression that adds `down`, `ydstogo`, and `yardline_100` to the combined model. Predict on the testing data and compare its testing Brier score with the three models above. Does the added complexity improve out-of-sample probability forecasts?

```{r challenge solution}
expanded_model <- glm(
  label_win ~ score_differential + posteam_spread + down + ydstogo + yardline_100,
  family = binomial,
  data = nfl_train
)

nfl_test <- nfl_test %>%
  mutate(
    expanded_prediction = predict(
      expanded_model,
      newdata = ., 
      type = "response"
    )
  )

nfl_test %>%
  summarize(Brier = mean((label_win - expanded_prediction)^2))
```
